### 2021-12-19 paper/main.tex
patched colorful language and transition to gh

\documentclass[12pt]{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with`a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{minted}
\usepackage{braket}
\usepackage{float}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{hypcap}

% xyqcirc stuff
\usepackage[frame,line,arrow,matrix,tips]{xy}	% all that is usually necessary

\doublespacing

\title{Quantum Bogosort}
\author{George Huebner}
\date{December 4, 2021}

\begin{document}
\maketitle

\begin{abstract}
Gag algorithms are a staple of computer science and provide many a good chuckle. Take, for instance, MiracleSort\textsuperscript{\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort the list:
\begin{minted}{python}
while isSorted(unsorted_list) is False:
    time.sleep(1000)
\end{minted}
Despite their apparent uselessness, joke algorithms provide valuable insight into algorithm design and complexity\textsuperscript{\cite{gruber_holzer_ruepp_2007}}. In this paper we propose an implementation to Quantum Bogosort, one such joke algorithm.
\end{abstract}

\section{Introduction}

Classic Bogosort is a joke sorting algorithm where a list is randomly permuted and checked to see if it is sorted. Essentially, in pseudocode:

\begin{minted}{python}
while isSorted(unsorted_list) is False:
    random.shuffle(unsorted_list)
\end{minted}

It has an average time complexity of $ O((n+1)!) $ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.

Quantum Bogosort\textsuperscript{\cite{ryder_2018}}, then, is a variation on classical bogosort. Based on the \href{https://www.pbs.org/wgbh/nova/manyworlds/pdf/dissertation.pdf}{many worlds} interpretation of quantum mechanics, QBS proposes that a quantum computer can superimpose all possible permutations of a list and destroy any universe in which a sorted list does not exist, thus sorting the list in $ O(1) $ time. Because parallel universes are non-communicating, there's no way to know the list is sorted other than by the principle of quantum necessity. The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) is essentially the same thing, but the quantum circuit must be run until the list is sorted.

Classical computers can already sort lists with a lower bound time complexity of $\Omega (N \log_2 N)$. Consequently, research into quantum sorting algorithms has been minimal, as the only optimizations possible are reducing space complexity. Because quantum computers do not have anywhere near enough qubits (let alone the error resilience) to realize this reduced complexity, sorting has been left to classical computers.

\section{Background}
Most people are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to test the QBS algorithm. Explaining the quantum mechanics that underlie this project is well beyond the scope of a math symposium, though, so let the physics enthusiast proceed carefully.

\subsection{Quantum Computers}
Quantum computers should be thought of not as processors that parse instructions, but rather circuits composed of gates that operate on a group of qubits (\textbf{qu}antum \textbf{bit}, the quantum unit of information). While revolutionary, quantum computers only achieve speedups in certain applications, and have not yet cleared a benchmark of true quantum supremacy due to current hardware implementations. Current noisy intermediate scale quantum (NISQ) computers are highly error prone and have comparatively few qubits and are better thought of as a separate processing unit to be orchestrated by a classical computer.

\subsection{Qubits}
Qubits have quantum properties of superposition, entanglement, and interference that distinguish them from classical bits. Whereas a classical bit is either 1 or 0, a qubit is in a superposition of 1 and 0, which are defined as the following kets:
$$ \ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \text{ (Classical 0)} $$
$$ \ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix} \text{ (Classical 1)} $$
Qubit superpositon can be thought of as having a certain amount (amplitude) of 'one-ness' and 'zero-ness' that can be represented as so:
$$ \ket{x} = \alpha \ket{0} + \beta \ket{1} \ | \ \text{ Probability}_{\ket{0}} = | \alpha |^2 \text{, } \text{ Probability}_{\ket{1}} = | \beta |^2 $$
%Note that $|x|$ here represents magnitude as $\alpha$ and $\beta$ are imaginary.
To actually make use of qubits, they must be measured, which collapses them to either $\ket{0}$ or $\ket{1}$. This is a probabilistic, quantum event, hence the probability\textsubscript{$\ket{x}$}. It is helpful to consider the \hyperref[fig:bsphere]{Bloch sphere} when analyzing qubit state and rotation around the Pauli axes; essentially, the 'North Pole' is defined as $\ket{0}$ and the 'South Pole' is defined as $\ket{1}$. The 'equator,' then, would be an even superposition of $\ket{0}$ and $\ket{1}$.

\begin{figure}[H]
    \centering
    \capstart
    \includegraphics[width=0.3\textwidth]{images/bloch_sphere.png}
    \caption{Bloch sphere\textsuperscript{\cite{wikipedia_2021}}}
    \label{fig:bsphere}
\end{figure}

\noindent Here are a few common qubit gates:
$$ \text{X}(\ket{0}) = \ket{1} $$
$$ \text{X}(\ket{1}) = \ket{0} $$
$$ \text{Z}(\ket{0}) = \ket{0} $$
$$ \text{Z}(\ket{1}) = -\ket{1} $$
$$ \text{H} = \text{R}_{\text{y}} (\frac{\pi}{2}) \text{Z} \ \ \ \text{(Hadamard gate)}$$
$$ \text{H}(\ket{0}) = \frac{1}{\sqrt{2}} \ket{0} + \frac{1}{\sqrt{2}} \ket{1} = \ket{+} $$
$$ \text{H}(\ket{1}) = \frac{1}{\sqrt{2}} \ket{0} - \frac{1}{\sqrt{2}} \ket{1} = \ket{-} $$
Notice that R\textsubscript{y} is an arbitrary rotation around the Pauli Y-axis; Similarly, R\textsubscript{x} and R\textsubscript{z} allow arbitrary rotations along the X- and Z-axes, respectively.\\
Multiple qubits can be used together for entanglement and amplitude interference, and a group of qubits is called a \textit{register}. Although a register can be simplified into a single tensor by taking the tensor product of the qubits in the register, keeping registers in kets makes the math much simpler. \\
Bra-ket (Dirac) notation is used as Kronecker products quickly become unwieldy when dealing with large qubit registers.

\begin{figure}[h]
    \centering
    $$ \ket{x} \otimes \ket{y} = \begin{bmatrix} x_0 \\ .. \\ x_n \end{bmatrix} \otimes \begin{bmatrix} y_0 \\ .. \\ y_m \end{bmatrix} = \begin{bmatrix} x_0 \cdot \begin{bmatrix} y_0 \\ .. \\ y_m \end{bmatrix} \\ .. \\ x_n \cdot \begin{bmatrix} y_0 \\ .. \\ y_m \end{bmatrix} \end{bmatrix} = \ket{xy} $$
    \caption{Elegance of Bra-ket notation}
    \label{fig:my_label}
\end{figure}

\subsection{Entanglement}
Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an 'if' clause in a classical programming language). This allows for multiple qubits to be entangled with each other, essentially tying the 'fate' of the qubits together so that they can no longer be thought of as individual qubits but rather a single system. Measuring one qubit will also collapse the other entangled qubits, even if they were never observed. A famous example of this is Schr\"odinger's cat. In this thought experiment, the quantum event of a radioactive substance decaying is entangled with the fate of the cat. If the cat dies, the radioactive substance must have decayed, and if the radioactive substance did not decay, the cat must be alive (or, at the very least, not killed by poison).

\section{Algorithm Design}
The implementation will be written using Qiskit, a Python framework for creating, running and simulating quantum circuits, and tested with the IBMQ Aer and Manila simulators.
Let us define the following:

\begin{itemize}
    \item A list of elements, $array$
    \item The length (cardinality) of this list, $l$
\end{itemize}

\noindent When we began designing this algorithm, two implementations were considered:
\begin{enumerate}
\item One approach would be to store the complete representation of the list in a qubit register. In Qiskit:
\begin{minted}{python}
array = [1, 2, 3]
size = 8
circuit = QuantumCircuit([QuantumRegister(size) for _ in range(len(array))])
\end{minted}
\textit{Note that Python numbers have dynamic size, but our 8-bit size would afford us an unsigned char, that is:}
$ x \in \mathbb{Z} \mid 0 \leq x < 256 $ \\
This clocks in at a respectable $ O(l \cdot s) $ space complexity, where $ s $ represents the size in bits of the data type in the list; While this method could potentially be useful in other quantum algorithms, this would essentially just transpose a classic algorithm onto a quantum circuit --- In the spirit of faithfully implementing QBS, we sought a more idiomatic quantum solution.

\item The other approach explored is essentially a quantum random number generator. Instead of storing the list, we create $ N = l! $ evenly balanced register states ($l!$ being the number of permutations of $array$). A set of states $ [0, N) \cap \mathbb{N} $ will be prepared, each with an even amplitude. Concisely:
    $$ \sum_{x=0}^{N-1} \frac{1}{\sqrt{N}} \ket{x_{BE}} $$
Note that the quantum register will always be represented in \href{https://en.wikipedia.org/wiki/Endianness}{big-endian}.

Because we're storing our number in binary, we want the power of 2 greater than or equal to $ l! $. This means we have $ O(\lceil \log_2 l! \rceil) $ space complexity, which seems scary at first. However, with present-day NISQ technology, this is significantly more favorable than $ O(l \cdot s) $ for any useful data type. Also keep in mind that because element size isn't a concern, any size element or even elements of different types could be stored in the list (as long as every type used can be compared to each other).

\begin{figure}[h]
    \centering
    \includegraphics[width=0.54\textwidth]{images/size_compare.png}
    \caption{Space complexity comparison}
    \label{fig:chart1}
\end{figure}

Before we proceed, we would like to discuss the naive approach to quantum RNG. Applying H\textsubscript{ALL} (apply Hadamard to all qubits) to the entire register can create a number of superimposed states $ n $ such that $ n $ is a power of 2. We could still use this output to generate a random number $ h $ such that $ h \in [i, j] $:
$$ h = i + \frac{\text{M}(register)}{2^{|register|}} \cdot j $$
%Where $|register|$ is the cardinality of the qubit register. \\
The simplicity of this approach is actually not a bad thing: Because quantum computers suffer from high incidences of soft errors (such as erroneous qubit flips), reducing the complexity of a quantum circuit is ideal. Despite this, we wanted the challenge of implementing an algorithm that creates an even superposition of any number of states.
\end{enumerate}

\section{Quantum Bogosort}
\subsection{QRNG Algorithm}
The main algorithm is recursive, essentially moving through the register from most to least significant qubit, performing a R\textsubscript{y} gate to split up a qubit into P 'parts,' and entangling the remaining qubits with controlled Hadamard (CH) gates. \\
Let's use $i = 77$ as an example. Recall that for storing an integer $i$, the most significant bit (MSB) used is $\lfloor \log_2 i \rfloor$: For example, the MSB of 77 would be $\lfloor \log_2 77 \rfloor = 64$. Because QRNG generates the integers $[0, 77)$ (excluding 77 itself in order to generate 77 integers), we know that $[0, 64)$ is also being generated, because $[0, 64) \subseteq [0, 77)$. \\
This means that every possible state of the qubit register with a leading zero (in this case, in the 64s place) should be generated. That's already really easy to do! Just use the leading zero as a control and apply Hadamard gates to the rest of the register. Here's a simple example:
\begin{align*}
& \ket{\psi} = \ket{0000000} \\
& H(\psi_0), \ \ \ \ \ \ \ket{\psi} = \frac{1}{\sqrt{2}} \ket{0000000} + \frac{1}{\sqrt{2}} \ket{1000000} \\
& CH(\psi_0, \psi_{[1, 6]}), \ X(\psi_0), \ \ \ \ \ \ \ \ket{\psi} = \frac{1}{\sqrt{128}}(\ket{0000000} + \ket{0000001} + ... + \ket{0111111}) + \frac{1}{\sqrt{2}} \ket{1000000}
\end{align*}
Note that CH is a singly controlled Hadamard applied to multiple qubits, not a multi-controlled Hadamard applied to one qubit. Also note the X gate applied to the MSB because we really want an anti-controlled, not a normal controlled gate. \\
The problem is that we've unevenly split the integer 77 into 2: $|[0, 64)| \neq |[64, 77)|$. Numbers in the range $[64, 77)$ have a $(\frac{1}{\sqrt{13}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured whereas numbers in the $[0, 64)$ range have a $(\frac{1}{\sqrt{64}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured. \\
To fix this, we have to change the $\frac{1}{\sqrt{2}}$ from the Hadamard gate to something that gives us a more complex superposition; to do this we use the arbitrary rotation gates. $X$ and $Y$ gates give us the desired rotations we want, but we chose $Y$ so we don't have to worry about phase (the sign of a qubit). Phase doesn't matter in this case because only the magnitudes effect the probability of measurement, but it might be important if using QRNG in another algorithm. \\
Superpositions can be represented using spherical coordinates. For example, on the \hyperref[fig:bsphere2]{Bloch sphere}, the spherical coordinates $(1, 0.955, 0)$ would represent the state $\sqrt{\frac{1}{3}} \ket{0} + \sqrt{\frac{2}{3}} \ket{1}$ because $\cos (0.955) \approx \sqrt{\frac{1}{3}}$. We can do simple algebra to get a $\theta$ for an arbitrary superposition. For QRNG, this looks like $\theta = 2\arccos(\sqrt{\frac{77 - MSB}{77}})$. \\
\begin{figure}[H]
    \centering
    \capstart
    \includegraphics[width=0.19\textwidth]{images/bloch_sphere2.png}
    \caption{Bloch sphere\textsuperscript{\cite{wikipedia_2021}}}
    \label{fig:bsphere2}
\end{figure}
\noindent Recall that we have to apply a $X$ gate because gates are controlled by $\ket{1}$, not $\ket{0}$. We can save an extra gate by doubling our rotation (hence the 2), which gives us $\sqrt{\frac{77 - MSB}{77}} \ket{0}$ instead, which flips right back to $\ket{1}$ with another $X$ gate. \\
\indent We then recursively repeat this process, adding the MSB as a control qubit to the R\textsubscript{y} and CH gates and using the 2nd most significant bit as the new MSB. Eventually we split all the states evenly and we're done. \\

\noindent The differences between what is conceptually happening and what is actually implemented are shown below. $\bullet$ is a control and $\circ$ is an anti-control (controls on $\ket{0}$ instead of $\ket{1}$). Note that H\textsubscript{ALL} is \textbf{not} a multi-qubit gate --- it's illustrated that way for clarity.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{images/circuits/conceptual_circuit.pdf}
    \caption{Conceptual circuit}
    \vspace{10mm}
    \includegraphics[width=\textwidth]{images/circuits/optimized_circuit.pdf}
    \caption{Circuit with proper controls}
    \label{fig:circs}
\end{figure}

\vspace{20mm}
\noindent Minor implementation details:
\begin{itemize}
    \item Multi-controlled Hadamard (MCH) gates are implemented manually with an ancilla qubit and multi-controlled X (MCX) gates. This could almost certainly be optimized more on a case-by-case basis (i.e. running \mintinline{python}{mode='noancilla'} for machines with fewer qubits).
    \item Especially when there are several qubits in the control register (and therefore the MCH gates are more costly), the MCX on the ancilla doesn't have to be reversed; instead, applying a  \mintinline{python}{'reset'} gate to reset the ancilla to 0 might be more optimal.
    \item \mintinline{python}{circuit} and \mintinline{python}{ancilla} don't change between recursions and needlessly add on to the stack frame of the function. Passing them as parameters is intended for clarity, but if you were to squeeze every drop of performance out they should be made global.
\end{itemize}
\subsubsection{Code}
\begin{minted}{python}
from qiskit import *
from math import acos, sqrt

def even_superpositon(circuit, register, controls, ancilla, P):
    if P <= pow(2, len(register)) / 2: # skip this iteration
        even_superpositon(circuit, register[1:], controls, ancilla, P)
        return
    if P == pow(2, len(register)): # reduced to H_ALL; base case
        if controls == []:
            for i in range(len(register)):
                circuit.h(register[i])
        elif len(register) > 0:
            circuit.mcx(controls, ancilla) # mcx to ancilla used in lieu of mch
            for i in range(len(register)):
                circuit.ch(ancilla, register[i])
            circuit.mcx(controls, ancilla)
        return

    extra_parts = P - pow(2, len(register) - 1)
    # this is the inverse of what we want, but saves us an X for ch
    theta = 2 * acos(sqrt(extra_parts / P))

    if controls == []:
        circuit.ry(theta, register[0])

        for i in range(1, len(register)):
            circuit.ch(register[0], register[i])
        controls += [register[0]]
    else:
        circuit.mcry(theta, controls, register[0])

        controls += [register[0]]
        circuit.mcx(controls + [], ancilla)
        for i in range(1, len(register)):
            circuit.ch(ancilla, register[i])
        circuit.mcx(controls, ancilla)

    circuit.x(register[0])

    even_superpositon(circuit, register[1:], controls, ancilla, extra_parts)
\end{minted}
\label{qrng}
\subsubsection{Testing}
Testing with an even superposition of 15 states:

\begin{minted}{python}
from math import ceil, log
from qiskit.visualization import plot_histogram

parts = 15
length = ceil(log(parts, 2))

register = QuantumRegister(length)
measurement = ClassicalRegister(length)
ancilla = QuantumRegister(1)
circuit = QuantumCircuit(register, ancilla, measurement)

even_superpositon(circuit, register, [], ancilla, parts)
circuit.measure(register, measurement)

count = execute(circuit, Aer.get_backend("aer_simulator"), shots=1024) //
.result().get_counts(circuit)
plot_histogram({key[::-1]: value for key, value in count.items()})
\end{minted}

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{images/example_circuit.png}
    \caption{Circuit diagram for 15 'parts'}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{images/example_measurement.png}
    \caption{Measurement results from Aer Simulator}
    \label{fig:measures}
\end{figure}

\subsection{Practicality Assessment}
Obviously bogosort is not meant to be a practical algorithm, but nevertheless we will conduct a review of the efficiency of QRNG.
\subsubsection{Time Complexity}
Although the original QBS proposal assumes the quantum circuit executes in $O(1)$ time, the depth (longest path) of the circuit does scale with input. The time a quantum circuit takes to execute is the sum of the gate execution times of the gates along the longest path of the circuit, and therefore time complexity scales with circuit depth. \\
For the integer $i$ passed into the QRNG algorithm, we allocate $\lceil \log_2 i \rceil + 1$ qubits. However, this algorithm is rather unique in that gate count and depth don't scale exactly with $i$, but rather \mintinline{python}{bin(i).count('1')}. That is, circuit depth scales with the number of ones in the binary representation of $i$. \\
Consequently, there are large drops in complexity around powers of 2:
\begin{align*}
&11111_2 \text{  (5 ones, high complexity)} \\ &+ 1 = 100000_2 \text{  (1 one, low complexity)}
\end{align*}
For example, $i = 256$ creates a circuit with depth 2 and width 17, but $i = 255$ has a circuit depth of 285 and a width of 17 (perhaps even more depending on transpilation target). This relationship is illustrated in \ref{fig:resources}.
% Also please note that depth + width $\geq$ total.

\begin{figure}[h]
    \centering
    \capstart
    \includegraphics[width=0.9\textwidth]{images/resources.png}
    \caption{Circuit depth, width, and total gate count in relation to $i$}
    \label{fig:resources}
\end{figure}
\vspace{10mm}
\noindent With $o = $ \mintinline{python}{bin(n).count('1')}, we have a time complexity of $O(o)$ for many-worlds and $O(o \cdot (n+1)!)$ for Copenhagen.

\subsubsection{Error Rate}
Of course, the main problem with this ballooning circuit size is the fidelity of the measurement; we work with imperfect machines and error rates can compound over time to make a measurement meaningless. 
That being said, this is a QRNG algorithm, so the noise and error generated don't make much of a difference in this scenario (given that the noise is truly random and evenly distributed) besides occasionally measuring a number $ \geq i$. \\
We attempted to create a script to measure the error rate of a circuit based on IBMQ benchmarks, however this only measures the probability of every gate succeeding. This doesn't take into account several errors cancelling each other out or other fortuitous errors, and consequently the measured success rate is astronomically small. A good extension on this would be to use the error rates to modify each gate: For example, an $X$ gate with error $= 1.9765\mathrm{e}{-4}$ would translate to R\textsubscript{x}($(1-1.9765\mathrm{e}{-4}) \cdot \pi$). This would allow for the approximation of the expected position on the Bloch sphere during measurement, thus providing a more accurate error rate.

\subsection{Bogosort}
This is a fairly straightforward implementation of Quantum Bogosort:
\begin{enumerate}
    \item Generate all permutations of list with length $l$ (symbolic representations are fine to avoid memory overhead)
    \item Use QRNG with parts = $l!$
    \item Using the measured value from step 2, randomly select a permutation of the list
    \item If the permutation is sorted, return
    \item Repeat steps 1-4
\end{enumerate}
\vspace{15mm}
Note that with the many-worlds interpretation we can stop at step 4, because a universe where the list is sorted must exist (assuming the list can be sorted) by quantum necessity. Perhaps disappointingly, there's no destroying of universes in this algorithm.

\subsection{Complete Algorithm}
Using QRNG algorithm defined in \ref{qrng}
\begin{minted}{python}
from itertools import permutations
from time import perf_counter

unsorted_list = [6, 10, 2589, 0, 47, 178, 324]
possibilities = list(permutations(unsorted_list))
unsorted_list.sort()

sorted = tuple(unsorted_list)

parts = len(possibilities)
length = ceil(log(parts, 2))
register = QuantumRegister(length)
measurement = ClassicalRegister(length)
ancilla = QuantumRegister(1)
circuit = QuantumCircuit(register, ancilla, measurement)

even_superpositon(circuit, register, [], ancilla, parts)
circuit.measure(register, measurement)

start = perf_counter()
while True:
    counts = execute(circuit, Aer.get_backend("aer_simulator"), shots=500) //
    .result().get_counts(circuit)

    max_count = 0
    index = None
    for num, count in counts.items():
        if count > max_count and (benum := int(num[::-1], 2)) < parts:
            max_count = count
            index = benum

    if index is not None:
        if possibilities[index] == sorted:
            break
end = perf_counter()

print(f"Finished Quantum Bogosort in {end-start} seconds on list of length //
      {len(unsorted_list)}")
\end{minted}
An impressive sort time of 258.02 seconds on a 7 element list... Quite bogus, indeed.

\section{Conclusion}
In this paper, we implemented the quantum bogosort algorithm using QRNG to randomly permute a list and check if it's sorted. Although QBS is a joke, the QRNG algorithm implemented could probably be useful in setting states where an H\textsubscript{ALL} wouldn't do the trick. Despite the fact that we couldn't find a similar QRNG algorithm, this seems too useful to be novel. \\
The QRNG algorithm presented could be further studied and improved for efficient preparation of arbitrarily superimposed qubits. Right now, QRNG generates an interval of integers, but this could be extended to generating an even superposition of arbitrary values. Additionally, as discussed in the practicality assessment, this algorithm does not scale well due to the prevalence of multi-controlled gates. \\
Another way to improve QBS would be to study how the list is sorted. Because the (non-unique) permutations of a list grows at $n!$, QRNG's $O(\log_2 n)$ space complexity quickly becomes overwhelmed with larger lists. Finding ways to reduce this inefficiency would be ideal, perhaps by using a 'divide and conquer' strategy (\`{a} la quicksort); However, this would defeat the purpose of bogosort! \\
We were able to both exercise previous knowledge of quantum computing and learn new skills in data collection and presentation. Additionally, we spent more time designing this algorithm because there's no previous implementation of QBS to compare against: Consequently we got to pursue many different solutions and choose the most interesting one. \\

\vspace{10mm}
\noindent All code was written and tested with Qiskit 0.32.0 \& Python 3.8 on WSL2 Ubuntu 20.04.1.

\vspace{15mm}
\noindent \emph{We acknowledge the use of IBM Quantum services for this work. The views expressed are those of the authors, and do not reflect the official policy or position of IBM or the IBM Quantum team.\textsuperscript{\cite{ibm_21}}}

\bibliographystyle{ieeetr}
\nocite{*}
\bibliography{sources}

\end{document}

### 2022-01-08 qbs.py
paper problems

@@ -1,0 +1,80 @@
+from qiskit import *
+from math import acos, sqrt, ceil, log
+from itertools import permutations
+from time import perf_counter
+
+def even_superpositon(circuit, register, controls, ancilla, P):
+    if P <= pow(2, len(register)) / 2: # skip this iteration
+        even_superpositon(circuit, register[1:], controls, ancilla, P)
+        return
+    if P == pow(2, len(register)): # reduced to H_ALL; base case
+        if controls == []:
+            for qb in register:
+                circuit.h(qb)
+        elif len(register) > 0:
+            circuit.mcx(controls, ancilla) # mcx to ancilla used in lieu of mch
+            for qb in register:
+                circuit.ch(ancilla, qb)
+            circuit.mcx(controls, ancilla)
+        return
+
+
+    extra_parts = P - pow(2, len(register) - 1)
+    # this is the inverse of what we want, but saves us an X for ch
+    theta = 2 * acos(sqrt(extra_parts / P))
+
+    if controls == []:
+        circuit.ry(theta, register[0])
+
+        for i in range(1, len(register)):
+            circuit.ch(register[0], register[i])
+        controls += [register[0]]
+    else:
+        circuit.mcry(theta, controls, register[0])
+
+        controls += [register[0]]
+        circuit.mcx(controls + [], ancilla)
+        for i in range(1, len(register)):
+            circuit.ch(ancilla, register[i])
+        circuit.mcx(controls, ancilla)
+
+    circuit.x(register[0])
+
+
+    even_superpositon(circuit, register[1:], controls, ancilla, extra_parts)
+
+
+if __name__ == "__main__":
+    set = [6, 10, 2589, 0, 47, 178, 324]
+    possibilities = list(permutations(set))
+    set.sort()
+
+    sorted = tuple(set)
+
+    parts = len(possibilities)
+    length = ceil(log(parts, 2))
+    register = QuantumRegister(length)
+    measurement = ClassicalRegister(length)
+    ancilla = QuantumRegister(1)
+    circuit = QuantumCircuit(register, ancilla, measurement)
+
+    even_superpositon(circuit, register, [], ancilla, parts)
+    circuit.measure(register, measurement)
+
+    start = perf_counter()
+    while True:
+        counts = execute(circuit, Aer.get_backend("aer_simulator"), shots=500).result().get_counts(circuit)
+
+        max_count = 0
+        index = None
+        for num, count in counts.items():
+            if count > max_count:
+                max_count = count
[…]

### 2022-01-08 resource_estimation.py

@@ -1,0 +1,119 @@
+from qiskit import QuantumCircuit, ClassicalRegister, QuantumRegister
+from qiskit import IBMQ
+from qiskit.compiler import transpile
+
+from time import perf_counter
+from math import ceil, log
+
+import qbs
+
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+def constructionTime(i):
+    length = ceil(log(i, 2))
+    register = QuantumRegister(length)
+    measurement = ClassicalRegister(length)
+    ancilla = QuantumRegister(1)
+    circuit = QuantumCircuit(register, ancilla, measurement)
+
+    #building
+    start = perf_counter()
+
+    qbs.even_superpositon(circuit, register, [], ancilla, i)
+    circuit.measure(register, measurement)
+
+    end = perf_counter()
+
+    print(f"Created QRNG circuit in {end-start} seconds.")
+    return circuit
+
+def compileTime(circuit, backend_name):
+    if not IBMQ.active_account():
+        global provider
+        provider = IBMQ.load_account()
+    backend = provider.get_backend(backend_name)
+
+    start = perf_counter()
+    new_circ = transpile(circuit, backend=backend, optimization_level=1)
+    end = perf_counter()
+
+    print(f"Transpiling QRNG for {backend_name} took {end - start} seconds.")
+    return new_circ
+
+
+def gateCount(circuit):
+    d = circuit.count_ops()
+    if "barrier" in d:
+        del d["barrier"]
+    if "snapshot" in d:
+        del d["snapshot"]
+
+    count = 0
+    tl = []
+    for k, v in d.items():
+        count += v
+        if k[0] == 'c':
+            tl.append((k, v))
+
+    print(f"Gate count: {count}")
+    if tl:
+        print(f"Additionally, there are ", end='')
+        it = iter(tl)
+        t = next(it)
+        if len(tl) == 1:
+            print(f"{t[1]} {t[0]} gates.")
+        else:
+            for k, v in it:
+                print(f"{v} {k} gates, ", end='')
+            print(f"& {t[1]} {t[0]} gates.")
+
+    return count
+    
+
+def metrics(i, sim_name):
+    print(f"Analyzing QRNG circuit with i = {i}.\n")
+
+    print("Metrics for original circuit:\n")
+    circ = constructionTime(i)
+    print(f"Depth: {circ.depth()}")
[…]

### 2022-03-05 paper/slides.tex
submission to city STEM fair

\documentclass{beamer}
\usetheme{metropolis}

\setbeamertemplate{bibliography item}[text]

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{braket}
\usepackage{float}
\usepackage{tikz}

\makeatletter
\newcommand\mathcircled[1]{%
  \mathpalette\@mathcircled{#1}%
}
\newcommand\@mathcircled[2]{%
  \tikz[baseline=(math.base), color=red] \node[draw,circle,inner sep=1pt] (math) {$\m@th#1#2$};%
}
\makeatother

\title{Quantum Bogosort}
\subtitle{Chicago STEM Fair, Design}
\date{March 18, 2022}
\author{George Huebner (Grade 12), Michael Caines (Teacher Sponsor)}
\institute{Walter Payton College Prep}
\begin{document}
  \maketitle

  \section{Introduction}

  \begin{frame}{Abstract}
    Gag algorithms are a staple of computer science and provide many a good chuckle. Take, for instance, MiracleSort\textsuperscript{\color{blue}\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort the list:
    \begin{center} \texttt{while isSorted(unsorted\_list) is False: \\ \hspace{-70} time.sleep(1000)} \end{center}
    Despite their apparent uselessness, joke algorithms provide valuable insight into algorithm design and complexity\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}}. In this paper we propose an implementation to Quantum Bogosort, one such joke algorithm.
  \end{frame}

  \begin{frame}{What is Quantum Bogosort?}
    \textit{Classical} Bogosort is a sorting algorithm.
    \begin{enumerate}
        \item Randomly permute the list.
        \item If the list is sorted, done! Otherwise go to step 1.
    \end{enumerate}
    This has unbounded worst-case time complexity and an expected time complexity of $O(n \cdot n!)$. \\
    \vspace{10}
    \textit{Quantum} Bogosort is very similar to classical Bogosort, except it uses a quantum computer to permute the list.
  \end{frame}

  \begin{frame}{Attachments}
  Code + Paper: \href{https://github.com/Borris-the-real-OG/Quantum_Bogosort}{\color{blue} \underline{github.com/Borris-the-real-OG/Quantum\_Bogosort}}
  \end{frame}

  \begin{frame}{Acknowledgements}
  \begin{itemize}
  \item \href{https://qiskit.org/}{\color{blue} \underline{Qiskit}}
  \item \emph{We acknowledge the use of IBM Quantum services for this work. The views expressed are those of the authors, and do not reflect the official policy or position of IBM or the IBM Quantum team.\textsuperscript{\color{blue}\cite{ibm_21}}}
  \item All code was written and tested with Qiskit 0.32.0 \& Python 3.8 on WSL2 Ubuntu 20.04.1.
  \item Special thanks to Joe Clapis
  \end{itemize}
  \end{frame}

  \begin{frame}{Purpose}
  Just like its classical counterpart, Quantum Bogosort is pretty useless. In fact, classical sorting algorithms have been proven to have a lower bound time complexity of $ \Omega (N \log_2 N) $: there's nothing to optimize! \\
  \textbf{\textit{However}}, QBS presents a good case study for quantum algorithms; we present an analysis of implementation and complexity of QBS in a similar vein to Gruber et al.'s\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}} analysis of classic bogosort.
  \end{frame}

  \section{Algorithm Design}

  \begin{frame}{QRNG}
    \begin{columns}[T,onlytextwidth]
        \column{0.48\textwidth}
            \vspace{20}
            Desired State: \\
            $$ \sum_{x=0}^{N-1} \frac{1}{\sqrt{N}} \ket{x_{BE}} $$
            This is quite easy to do with powers of 2; we just need to use Hadamard gates with control qubits to chunk a big number into smaller subproblems.
       \column{0.48\textwidth}
           \begin{align*}
                & \ket{\psi} = \mathcircled{\color{black} \frac{1}{\sqrt{2}}} \ket{0000} + \frac{1}{\sqrt{2}} \ket{1000} \\
                & \mathbf{CH(\psi_0, \psi_{[1, 3]}), \ X(\psi_0)} \\
                & \ket{\psi} = \mathcircled{\color{black} \frac{1}{\sqrt{16}}}(\ket{0000} + \ket{0001} + \\ 
                & ... + \ket{0111}) + \frac{1}{\sqrt{2}} \ket{1000}
            \end{align*}
            \color{red} We mostly care about amplitude as opposed to register state
    \end{columns}
  \end{frame}

  \begin{frame}{QRNG}
    \textbf{Limitation:} Hadamard gates create the state $ \frac{1}{\sqrt{2}} \ket{0} + \frac{1}{\sqrt{2}} \ket{1} $. What about states that aren't powers of 2? \\ \\
    \textbf{Solution:} Arbitrary rotation gates allow for qubit superposition with arbitrary amplitudes. \\
    Hadamard is just a rotation of $ \frac{\pi}{2} $ around the Pauli Y-axis!
    \begin{align*}
        &\ket{\psi} = \ket{0} \\
        & \text{R}_{\text{y}}(\theta) \ket{\psi} = \begin{bmatrix} \cos \frac{\theta}{2} & -\sin \frac{\theta}{2} \\ \sin \frac{\theta}{2} & \cos \frac{\theta}{2} \end{bmatrix} \ket{\psi} = -\sin \frac{\theta}{2} \ket{0} + \cos \frac{\theta}{2} \ket{1}
    \end{align*}
    \textit{Everything in this project is measured in Pauli-Z basis, so we don't worry about phase or imaginary numbers.}
  \end{frame}

  \begin{frame}{Algorithm}
  \begin{enumerate}
      \item If `parts' $= 2^\texttt{len(register)}$, use the control register to apply multi-controlled Hadamard gates to the rest of the register. Done.
      \item If `parts' $\leq 2^\texttt{len(register)}-1$, add \texttt{register[0]} to the control register and skip to step 5.
      \item Calculate $\theta$, R\textsubscript{y}$(\theta)$ the first register qubit, and add it to the control register.
      \item Using the control register, apply multi-controlled Hadamard gates to the rest of the register.
      \item Recurse on the remaining qubits. \\
  \end{enumerate}
  \end{frame}

  \begin{frame}{Worked Example}
    \begin{columns}
        \column{0.5\textwidth}
        $$i = 77, \ \ket{\psi} = \ket{0000000}$$
        \column{0.5\textwidth}
        $$\theta_0 = 2\arccos \sqrt{\frac{77-\color{violet}64}{77}}$$
        \color{violet} Most Significant Bit in register
    \end{columns}
    \begin{align*}
      &\text{R}_y(\theta_0, \psi_0), \ \color{lightgray}\text{X}(\psi_0)\color{black}, \ \text{CH}(\psi_0, \psi_{[1,6]}), \ \text{X}(\psi_0) = \\
      &\color{green} \frac{1}{\sqrt{77}} (\ket{0000000} + \ket{0000001} + ... + \ket{0111111}) \color{black} + \color{red} \sqrt{\frac{13}{77}} \ket{1000000}
      \end{align*}
      \hspace{60} \color{green} Done with this part \hspace{70} \color{red} New subproblem
  \end{frame}

  \section{Practicality Assessment/Analysis}

  \begin{frame}{Circuit}
    \begin{figure}
        \centering
        \begin{tikzpicture}[lablum/.style 2 args={label=below:#1 #2,name=img-#1},
        marr/.style={line width=1mm,-latex}]
         \matrix[column sep=1cm,row sep=5mm] (mat)
         { \node[lablum={1}{Conceptual Circuit\textsuperscript{\color{blue}\cite{chuang}}}]{\includegraphics[width=0.8\textwidth]{images/circuits/conceptual_circuit.pdf}};\\ \\
         \node[lablum={2}{Compiled to Aer Simulator}]{\includegraphics[width=0.8\textwidth]{images/example_circuit.png}};\\};
         \draw[marr] ([yshift=-7mm]img-1.south) coordinate (aux) 
         -- (img-2.north-|aux);
        \end{tikzpicture}
    \end{figure}
   \end{frame}

   \begin{frame}{Results}
       \begin{columns}
           \column{0.48\textwidth}
           \begin{center}
           \includegraphics[width=\linewidth]{images/example_measurement.png}
           \textit{Example measurement with }\texttt{parts = 15}. \\
           \end{center}
           \textbf{Complexity} \\
           $n =$ \# elements, $o =$ \texttt{bin(n).count(`1')} \\
           Many-worlds: $O(o)$ \\
           Copenhagen: $O(o \cdot n!)$

           \column{0.48\textwidth}
           Gate count scales with the number of `1's in the binary representation, \textit{\textbf{not}} the size of the number.
           \includegraphics[width=\linewidth]{images/resources.png}
           \textit{Circuit depth, width, and total gate count in relation to} $ i $.
       \end{columns}
   \end{frame}

  \begin{frame}{Conclusion}
  Quantum Bogosort is not a useful algorithm, but it's helpful in teaching quantum computing fundamentals in a fun way. \\ \\
  Although infeasible due to its ballooning gate count, the QRNG algorithm could be repurposed for quickly preparing evenly balanced states among qubits. \\ \\
  By trying to optimize around the abysmal $ O(\log_2 n!) $ space complexity via `dividing and conquering', we end up implementing quicksort.
  \end{frame}

  \begin{frame}{Bibliography}
  \bibliographystyle{ieeetr}
  \bibliography{sources}
  \end{frame}

\end{document}

### 2022-03-05 qbs.py

@@ -44,12 +44,21 @@ def even_superpositon(circuit, register, controls, ancilla, P):
     even_superpositon(circuit, register[1:], controls, ancilla, extra_parts)
 
 
+def isSorted(ls):
+    if not ls:
+        return True
+
+    prev = ls[0]
+    for i in ls:
+        if i < prev:
+            return False
+        prev = i
+    return True
+
+
 if __name__ == "__main__":
     set = [6, 10, 2589, 0, 47, 178, 324]
     possibilities = list(permutations(set))
-    set.sort()
-
-    sorted = tuple(set)
 
     parts = len(possibilities)
     length = ceil(log(parts, 2))
@@ -73,7 +82,7 @@ if __name__ == "__main__":
                 index = int(num[::-1], 2)
 
         if index is not None:
-            if possibilities[index] == sorted:
+            if isSorted(possibilities[index]):
                 break
     end = perf_counter()
 

### 2022-03-08 paper/images/abstract.tex
let's hope I didn't just dox myself

% this is SO DUMB
\documentclass{article}
\usepackage{multicol}

\begin{document}
\begin{multicols}{3}
    \centering
    \textbf{Purpose} \\
    \raggedright To explore quantum computing and development of NISQ devices by implementing the theoretical ``quantum bogosort'' gag algorithm. \\
    \columnbreak
    \centering
    \textbf{Procedure} \\
    \raggedright Implemented with Qiskit and tested on Aer simulator and IBMQ Lima (5 qubit machine). \\
    \columnbreak
    \centering
    \textbf{Conclusion} \\
    \raggedright Quantum bogosort is a funky way to introduce quantum computing fundamentals to beginners. Novel QRNG algorithm provides insight into crafting arbitrary qubit states but is not practically useful.
\end{multicols}
\end{document}

### 2022-03-08 paper/main.tex
let's hope I didn't just dox myself

@@
 \usepackage{hypcap}
+\usepackage{fancyhdr}
 
@@
 \author{George Huebner}
-\date{December 4, 2021}
+\date{March 18, 2022}
 
 \begin{document}
+\pagenumbering{gobble}
+
+\begin{titlepage}
 \maketitle
+\centering
+\large Chicago STEM Fair, Computer Science Research, Design Category \\
+\vspace{3em}
+\huge Sponsor \\
+\vspace{1em}
+\large Michael Caines \hspace{15em}
+\end{titlepage}
+
+{
+\hypersetup{hidelinks}
+\tableofcontents
+}
+\pagebreak
+
+\pagestyle{fancy}
+\fancyhead[R]{\textit{Quantum Bogosort}, Huebner}
+\fancyhead[L]{}
+
+\pagenumbering{arabic}
 
 \begin{abstract}
-Gag algorithms are a staple of computer science and provide many a good chuckle.
-Take, for instance, MiracleSort\textsuperscript{\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort the list:
+\noindent Gag algorithms are a staple of computer science and provide many a good chuckle.
+Take, for instance, MiracleSort\textsuperscript{\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort a list:
 \begin{minted}{python}
@@
 
-It has an average time complexity of $ O((n+1)!) $ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.
+\noindent It has an average time complexity of $ O((n+1)!) $ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.
 
@@
 \vspace{10mm}
-\noindent With $o = $ \mintinline{python}{bin(n).count('1')}, we have a time complexity of $O(o)$ for many-worlds and $O(o \cdot (n+1)!)$ for Copenhagen.
+\noindent With $o = $ \mintinline{python}{bin(n).count('1')}, we have a time complexity of $O(o)$ for many-worlds and $O(o \cdot n!)$ for Copenhagen.
 
@@
 
+def isSorted(ls):
+    if not ls:
+        return True
+
+    prev = ls[0]
+    for i in ls:
+        if i < prev:
+            return False
+        prev = i
+    return True
+
 unsorted_list = [6, 10, 2589, 0, 47, 178, 324]
 possibilities = list(permutations(unsorted_list))
-unsorted_list.sort()
-
-sorted = tuple(unsorted_list)
 
@@
     if index is not None:
-        if possibilities[index] == sorted:
+        if isSorted(possibilities[index]):
             break

### 2022-09-06 paper/main.tex
don't steal my stuff until ISEF is over 😭
major post-ISEF cleanup
transitioned to Git LFS hosting of images/PDFs, added ISEF materials,
and updated LaTeX source. I think testing.ipynb has some changes on a
different machine from when I added density matrix calculations...

@@
 \title{Quantum Bogosort}
-\author{George Huebner}
-\date{March 18, 2022}
+\author{George Huebner}% (Grade 12), Michael Caines (Teacher sponsor)}
+\date{May 7, 2022}
 
 \begin{document}
+\renewcommand{\abstractname}{Preface}
 \pagenumbering{gobble}
@@
 \centering
-\large Chicago STEM Fair, Computer Science Research, Design Category \\
-\vspace{3em}
-\huge Sponsor \\
-\vspace{1em}
-\large Michael Caines \hspace{15em}
+\large Walter Payton College Prep, Chicago IL, United States \\
+ISEF Systems Software/Algorithms (SOFT055)% \\
+%\large Illinois STEM Fair, Computer Science Research, Design Category \\
+%\vspace{3em}
+%\huge Sponsor \\
+%\vspace{1em}
+%\large Michael Caines \hspace{15em}
 \end{titlepage}
+
+\pagestyle{fancy}
+\fancyhead[R]{Huebner: \textit{Quantum Bogosort} (SOFT055)}
+\fancyhead[L]{}
 
@@
 
-\pagestyle{fancy}
-\fancyhead[R]{\textit{Quantum Bogosort}, Huebner}
-\fancyhead[L]{}
-
 \pagenumbering{arabic}
@@
 \begin{abstract}
-\noindent Gag algorithms are a staple of computer science and provide many a good chuckle.
+\noindent Joke algorithms are a staple of computer science and provide many a good chuckle.
 Take, for instance, MiracleSort\textsuperscript{\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort a list:
@@
 
-Classic Bogosort is a joke sorting algorithm where a list is randomly permuted and checked to see if it is sorted. Essentially, in pseudocode:
+Classic Bogosort is a joke sorting algorithm where a list is randomly permuted and checked to see if it is sorted. In pseudocode:
 
@@
 
-\noindent It has an average time complexity of $ O((n+1)!) $ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.
+\noindent It has an average time complexity of $O((n+1)!)$ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.
 
-Quantum Bogosort\textsuperscript{\cite{ryder_2018}}, then, is a variation on classical bogosort.
+Quantum Bogosort\textsuperscript{\cite{the_other_tree_2009}} (QBS), then, is a variation on classical bogosort.
 Based on the \href{https://www.pbs.org/wgbh/nova/manyworlds/pdf/dissertation.pdf}{many worlds} interpretation of quantum mechanics, QBS proposes that a quantum computer can superimpose all possible permutations of a list and destroy any universe in which a sorted list does not exist, thus sorting the list in $ O(1) $ time.
 Because parallel universes are non-communicating, there's no way to know the list is sorted other than by the principle of quantum necessity.
-The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) is essentially the same thing, but the quantum circuit must be run until the list is sorted.
+The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) uses the same algorithm, but the quantum circuit must be run until the list is sorted.
 
@@
 \section{Background}
-Most people are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to test the QBS algorithm.
-Explaining the quantum mechanics that underlie this project is well beyond the scope of a math symposium, though, so let the physics enthusiast proceed carefully.
+Most people are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to develop and test QBS.
+This is not meant to be a complete explaination of the quantum mechanics that underlie this project, though, so let the physics enthusiast proceed carefully.
 
@@
 Current noisy intermediate scale quantum (NISQ) computers are highly error prone and have comparatively few qubits and are better thought of as a separate processing unit to be orchestrated by a classical computer.
-
 \subsection{Qubits}
@@
 $$ \ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix} \text{ (Classical 1)} $$
-Qubit superpositon can be thought of as having a certain amount (amplitude) of 'one-ness' and 'zero-ness' that can be represented as so:
+Qubit superpositon can be thought of as having a certain amount (amplitude) of `one-ness' and `zero-ness' that can be represented as so:
 $$ \ket{x} = \alpha \ket{0} + \beta \ket{1} \ | \ \text{ Probability}_{\ket{0}} = | \alpha |^2 \text{, } \text{ Probability}_{\ket{1}} = | \beta |^2 $$
@@
 This is a probabilistic, quantum event, hence the probability\textsubscript{$\ket{x}$}.
-It is helpful to consider the \hyperref[fig:bsphere]{Bloch sphere} when analyzing qubit state and rotation around the Pauli axes; essentially, the 'North Pole' is defined as $\ket{0}$ and the 'South Pole' is defined as $\ket{1}$.
-The 'equator,' then, would be an even superposition of $\ket{0}$ and $\ket{1}$.
+It is helpful to consider the \hyperref[fig:bsphere]{Bloch sphere} when analyzing qubit state and rotation around the Pauli axes; essentially, the `North Pole' is defined as $\ket{0}$ and the `South Pole' is defined as $\ket{1}$.
+The `equator,' then, would be an even superposition of $\ket{0}$ and $\ket{1}$.
 
@@
 Although a register can be simplified into a single tensor by taking the tensor product of the qubits in the register, keeping registers in kets makes the math much simpler.
-\\
 Bra-ket (Dirac) notation is used as Kronecker products quickly become unwieldy when dealing with large qubit registers.
@@
 \subsection{Entanglement}
-Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an 'if' clause in a classical programming language).
+Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an `if' clause in a classical programming language).
 This allows for multiple qubits to be entangled with each other, essentially tying the 'fate' of the qubits together so that they can no longer be thought of as individual qubits but rather a single system.
@@
 \item The other approach explored is essentially a quantum random number generator.
-Instead of storing the list, we create $ N = l!
-$ evenly balanced register states ($l!$ being the number of permutations of $array$).
-A set of states $ [0, N) \cap \mathbb{N} $ will be prepared, each with an even amplitude.
+Instead of storing the list, we create $N = l!$ evenly balanced register states ($l!$ being the number of permutations of $array$).
+A set of states $[0, N) \cap \mathbb{N}$ will be prepared, each with an even amplitude.
 Concisely:
@@
 
-Because we're storing our number in binary, we want the power of 2 greater than or equal to $ l!
-$.
-This means we have $ O(\lceil \log_2 l!
-\rceil) $ space complexity, which seems scary at first.
-However, with present-day NISQ technology, this is significantly more favorable than $ O(l \cdot s) $ for any useful data type.
+Because we're storing our number in binary, we want the power of 2 greater than or equal to $l!$.
+This means we have $O(\lceil \log_2 l!
+\rceil)$ space complexity, which seems scary at first.
+However, with present-day NISQ technology, this is significantly more favorable than $O(l \cdot s)$ for any useful data type.
 Also keep in mind that because element size isn't a concern, any size element or even elements of different types could be stored in the list (as long as every type used can be compared to each other).
@@
 \section{Quantum Bogosort}
-\subsection{QRNG Algorithm}
+\subsection{Quantum Random Number Generation (QRNG) Algorithm}
 The main algorithm is recursive, essentially moving through the register from most to least significant qubit, performing a R\textsubscript{y} gate to split up a qubit into P 'parts,' and entangling the remaining qubits with controlled Hadamard (CH) gates.
@@
 Just use the leading zero as a control and apply Hadamard gates to the rest of the register.
-Here's a simple example:
+Here's an example:
 \begin{align*}
@@
 For example, on the \hyperref[fig:bsphere2]{Bloch sphere}, the spherical coordinates $(1, 0.955, 0)$ would represent the state $\sqrt{\frac{1}{3}} \ket{0} + \sqrt{\frac{2}{3}} \ket{1}$ because $\cos (0.955) \approx \sqrt{\frac{1}{3}}$.
-We can do simple algebra to get a $\theta$ for an arbitrary superposition.
+We can do some algebra to get a $\theta$ for an arbitrary superposition.
 For QRNG, this looks like $\theta = 2\arccos(\sqrt{\frac{77 - MSB}{77}})$.
@@
     \includegraphics[width=0.9\textwidth]{images/circuits/conceptual_circuit.pdf}
-    \caption{Conceptual circuit}
+    \caption{Conceptual circuit\textsuperscript{\cite{chuang}}}
     \vspace{10mm}
     \includegraphics[width=\textwidth]{images/circuits/optimized_circuit.pdf}
-    \caption{Circuit with proper controls}
+    \caption{Circuit with proper controls\textsuperscript{\cite{chuang}}}
     \label{fig:circs}
@@
 
-\vspace{20mm}
+\vspace{10mm}
 \noindent Minor implementation details:
@@
     \includegraphics[width=\textwidth]{images/example_circuit.png}
-    \caption{Circuit diagram for 15 'parts'}
+    \caption{Circuit diagram for 15 `parts'}
+    \label{fig:circ_diagram}
 \end{figure}
@@
 Obviously bogosort is not meant to be a practical algorithm, but nevertheless we will conduct a review of the efficiency of QRNG.
-\subsubsection{Time Complexity}
+\subsubsection{Complexity}
 Although the original QBS proposal assumes the quantum circuit executes in $O(1)$ time, the depth (longest path) of the circuit does scale with input.
@@
 \\
-We attempted to create a script to measure the error rate of a circuit based on IBMQ benchmarks, however this only measures the probability of every gate succeeding.
+%We attempted to create a script to measure the error rate of a circuit based on IBMQ benchmarks, however this only measures the probability of every gate succeeding.
 This doesn't take into account several errors cancelling each other out or other fortuitous errors, and consequently the measured success rate is astronomically small.
@@
 This would allow for the approximation of the expected position on the Bloch sphere during measurement, thus providing a more accurate error rate.
+One way we can model error is by wrapping our gates with error gates --- identity gates with an error rate corresponding to a noise profile sampled from real quantum hardware.
+This approach works with gate errors, including measurement and reset errors, and could also be applied to decoherence and relaxation errors.
+\\
+Often times noise on hardware will turn a pure quantum state into a mixed state, a composition of pure states with classical probabilities (which is \textbf{not} the same as superposition); for example, perhaps an R\textsubscript{Y}$(\frac{\pi}{2})$ will succeed $80\%$ of the time and over/undershoot by $\frac{\pi}{6}$ 20\% of the time.
+We can write this mixed state as
+\begin{align*}
+\ket{\psi_0} &= \frac{1}{\sqrt{2}} \left( \ket{0} + \ket{1} \right) \ \ \ \text{(80\% chance)} \\
+\ket{\psi_1} &= \frac{1}{\sqrt{4}} \ket{0} + \sqrt{\frac{3}{4}} \ket{1} \ \ \ \text{(10\% chance)} \\
+\ket{\psi_2} &= \sqrt{\frac{3}{4}} \ket{0} + \frac{1}{\sqrt{4}} \ket{1} \ \ \ \text{(10\% chance)}
+\end{align*}
+Density matrices are particularly convenient because we can represent this mixed state in a single matrix:
+\begin{align*}
+    \rho &= \frac{4}{5} \ket{\psi_0}\bra{\psi_0} + \frac{1}{10} \ket{\psi_1}\bra{\psi_1} + \frac{1}{10} \ket{\psi_2}\bra{\psi_2} \\
+    &= \frac{4}{5} \begin{bmatrix} \frac{1}{2} & \frac{1}{2} \\ \frac{1}{2} & \frac{1}{2} \end{bmatrix} + \frac{1}{10} \begin{bmatrix} \frac{1}{4} & \frac{\sqrt{3}}{4} \\ \frac{\sqrt{3}}{4} & \frac{3}{4} \\ \end{bmatrix} + \frac{1}{10} \begin{bmatrix} \frac{3}{4} & \frac{\sqrt{3}}{4} \\ \frac{\sqrt{3}}{4} & \frac{1}{4} \\ \end{bmatrix} \\
+    &= \begin{bmatrix} \frac{1}{2} & \frac{2}{5} + \frac{\sqrt{3}}{20} \\ \frac{2}{5} + \frac{\sqrt{3}}{20} & \frac{1}{2} \\ \end{bmatrix} \ \ \text{Purity: 0.9736}
+\end{align*}
+In the case of quantum bogosort, we can calculate the density matrix for the idealized ($\rho$) and noisy ($\rho'$) circuits; Both are transpiled for IBMQ Santiago and the latter is injected with Santiago's noise profile.
+$\rho'$'s purity of 0.42118 reflects that noise heavily affects the fidelity of the qubits.
+\begin{gather*}
+    \rho = \begin{bmatrix}
+        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
+        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
+        \vdots & \vdots & \ddots & \vdots & \vdots \\
+        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
+        0 & 0 & \cdots & 0 & 0 \\
+    \end{bmatrix} \\
+    \rho' = \begin{bmatrix}
+        0.06997 & 0.04275 - 0.00057i & \cdots & 0.0376 + 0.00028i & 0.00036 + 0.00024i  \\
+        0.04275 + 0.00057i & 0.06646 & \cdots & 0.03687 + 0.00025i & 0.00002 + 0.0001i  \\
+        \vdots & \vdots & \ddots & \vdots & \vdots \\
+        0.0376 - 0.00028i & 0.03687 - 0.00025i & \cdots & 0.06272 & 0.00217 + 0.00045i  \\
+        0.00036 - 0.00024i & 0.00002 - 0.0001i & \cdots & 0.00217 - 0.00045i & 0.01268  \\
+    \end{bmatrix}
+\end{gather*}
 
@@
 
+\pagebreak
+
 \section{Conclusion}
-In this paper, we implemented the quantum bogosort algorithm using QRNG to randomly permute a list and check if it's sorted.
-Although QBS is a joke, the QRNG algorithm implemented could probably be useful in setting states where an H\textsubscript{ALL} wouldn't do the trick.
-Despite the fact that we couldn't find a similar QRNG algorithm, this seems too useful to be novel.
+In this paper, we implemented the quantum bogosort algorithm using a novel QRNG to randomly permute the list.
+Although QBS is a joke, the QRNG algorithm implemented could potentially be useful in setting states where an H\textsubscript{ALL} wouldn't do the trick.
 \\
@@
 Another way to improve QBS would be to study how the list is sorted.
-Because the (non-unique) permutations of a list grows at $n!$, QRNG's $O(\log_2 n)$ space complexity quickly becomes overwhelmed with larger lists.
-Finding ways to reduce this inefficiency would be ideal, perhaps by using a 'divide and conquer' strategy (\`{a} la quicksort); However, this would defeat the purpose of bogosort!
+Because the (non-unique) permutations of a list grows at $n!$, QRNG's $O(\log_2 n!)$ space complexity quickly becomes overwhelmed with larger lists.
+Finding ways to reduce this inefficiency would be ideal, perhaps by using a `divide and conquer' strategy (\`{a} la quicksort); However, this would defeat the purpose of bogosort!
 \\
-We were able to both exercise previous knowledge of quantum computing and learn new skills in data collection and presentation.
+Although this is one way to achieve even measurement amplitudes over a register, it's not necessarily the \textit{fastest} way; transpilation optimizations and optimal gate decompositions pose interesting extensions to this problem that are especially pertinent with current quantum computers.
+\\
+%We were able to both exercise previous knowledge of quantum computing and learn new skills in data collection and presentation.
 Additionally, we spent more time designing this algorithm because there's no previous implementation of QBS to compare against: Consequently we got to pursue many different solutions and choose the most interesting one.
@@
 
+\pagebreak
+
 \bibliographystyle{ieeetr}

### 2022-09-06 paper/research_plan.tex
major post-ISEF cleanup
transitioned to Git LFS hosting of images/PDFs, added ISEF materials,
and updated LaTeX source. I think testing.ipynb has some changes on a
different machine from when I added density matrix calculations...

\documentclass[12pt]{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{fancyhdr}
%\renewcommand{\headrulewidth}{0pt}
\usepackage{blindtext}
\usepackage{minted}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{braket}

\usepackage{etoolbox}
\patchcmd{\thebibliography}{\section*{\refname}}{}{}{}

% Set page size and margins
% Replace `letterpaper' with`a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

\begin{document}

\pagestyle{fancy}
\fancyhead[L]{George Huebner, Grade 12 \\ Walter Payton College Prep}
\fancyhead[R]{}

\centering \null
\vspace{2em}
\Huge ISEF Research Plan \\
\vspace{2em}
\raggedright \normalsize

\section{Project Rationale}
Quantum computers provide exciting new possibilities in the field of computer science --- superpolynomial speedups, quantum machine learning, and chemistry simulations are among the many promising breakthroughs quantum computers can achieve. Now is a particularly exciting time in the field of quantum information theory because hardware is finally catching up to decades of theory.
%--- the first undisputed demonstration of quantum supremacy (solving a problem that a classical computer cannot in a useful amount of time) was achieved in 2020.
But for all the hype surrounding quantum computing, relatively few programmers know how to harness their power. \par% many find the advanced mathematics and physics surrounding the subject to be difficult and off-putting. \\

However, just as one doesn't need to fully understand the nuances of silicon wafer manufacturing in order to program a classical computer, programmers do not need to fully understand the underlying quantum mechanics in order to program a quantum computer. Knowing even a small amount of quantum information theory can enable researchers to utilize the power of quantum computers. Therein lies the goal of the ``quantum bogosort'' algorithm: to spread awareness of quantum computing with a quirky real-world algorithm that beginners can understand and appreciate.

%Isaac Chuang's ``Quantum Computation and Quantum Information'' provides a more

%by creating a non-trivial algorithm that teaches basis gates, basic qubit manipulation, superposition and entanglement
%the goal of this project is to help teach quantum computing fundamentals in a way that engages

\section{Engineering Goal}
Implement the joke ``quantum bogosort'' algorithm using a non-trivial circuit that covers most fundamental quantum computing concepts. The algorithm should be sufficiently complex that it presents an interesting challenge to beginners, but not so difficult that it becomes incomprehensible. The reason a joke algorithm was chosen instead of an integer factorization or quantum error correction routine was to create something novel and, most importantly, fun! \par

Because even relatively simple quantum algorithms (i.e. Deutsch-Jozsa) can be challenging for beginners, our algorithm will be designed exclusively around basic gates.
Additionally, although it would be possible to create a trivial QRNG (\textbf{q}uantum \textbf{r}andom \textbf{n}umber \textbf{g}enerator) with a probability $\frac{1}{2^{n}} \ | \ n \in \mathbb{Z}$, this doesn't accomplish the stated goal of teaching fundamentals because it only uses one gate.

%Notably, a key tenet that is not covered by the quantum bogosort algorithm is amplitude interference. To rectify this, a more general algorithm was created:
%We can rewrite any list (sorted or not) as a dictionary where each unique element is a key with value equal to the number of its occurrences. For example, the list $[4, 6, 7, 4, 8, 8]$ could be rewritten as $\{4{:} \ 2, 6{:} \ 1, 7{:} \ 1, 8{:} \ 2 \}$. 
%$$ \sum_{x=0}^{N-1} \frac{1}{\sqrt{N}} \ket{x_{BE}} $$

\section{Procedure}

%\subsection{Safety}
%\blindtext

\subsection{Experimentation}
\begin{enumerate}
    \item Research how to integrate quantum into classical bogosort and craft desired qubit state
    \item Design conceptual circuit and classical list processing routine
    \item Implement conceptual circuit (using Qiskit)
    \item Continue to iterate on design, using quantum simulators for testing
    \item Resources permitting, collect data on real quantum hardware (using IBM Quantum)
\end{enumerate}

\subsection{Data Analysis}
The correctness of the algorithm can be verified by analyzing the amplitudes of statevectors for a given circuit, because circuits are ultimately just compositions of matrices. Consequently, statistical analysis to determine if results are due to chance is not required. This is also why simulators generally suffice in the event quantum hardware is inaccessible --- choosing an output state with a classically generated random number may not be truly random, but it's certainly random \textit{enough}.  \\
The random number generated by the quantum circuit is used to select one of the permutations of the list, which will then be checked to see if it is sorted. Note that the implementation of \href{https://docs.python.org/3/library/itertools.html?highlight=permutation#itertools.permutations}{\mintinline{python}{permutations()}} can be arbitrary as long as it is a pure function --- that is, for a given list, the order of the permutations doesn't matter, but it must be consistent.

%\section{Project Summary}
%\blindtext

\section{Bibliography}

\bibliographystyle{ieeetr} % abbrv or apalike
\nocite{*}
\bibliography{sources}

\end{document}

### 2022-09-06 paper/slides.tex
let's hope I didn't just dox myself
major post-ISEF cleanup
transitioned to Git LFS hosting of images/PDFs, added ISEF materials,
and updated LaTeX source. I think testing.ipynb has some changes on a
different machine from when I added density matrix calculations...

@@
 \usepackage{graphicx}
+\usepackage{graphics}
 \usepackage{setspace}
@@
 \usepackage{float}
+\usepackage{nicematrix}
 \usepackage{tikz}
+\usetikzlibrary{fit}
 
@@
 
+\newcommand\scalemath[2]{\scalebox{#1}{\mbox{\ensuremath{\displaystyle #2}}}}
+
 \title{Quantum Bogosort}
-\subtitle{Chicago STEM Fair, Design}
-\date{March 18, 2022}
+\subtitle{Systems Software/Algorithms}
+\date{May 7, 2022}
 \author{George Huebner (Grade 12), Michael Caines (Teacher Sponsor)}
-\institute{Walter Payton College Prep}
+\institute{Walter Payton College Prep, Chicago IL, United States}
 \begin{document}
+  \begin{frame}
     \maketitle
+    \begin{tikzpicture}[overlay, remember picture]
+     \node[xshift=-1cm,yshift=-0.33cm, draw=blue] at (current page.north east) {SOFT055};
+     \end{tikzpicture}
+  \end{frame}
 
-  \section{Introduction}
+  % we looove red tape
+  %\begin{frame}[plain, noframenumbering]{Attachments}
+    %\begin{columns}
+        %\column{0.5\textwidth}
+            %\includegraphics[width=\linewidth]{images/abstract.png}
+        %\column{0.5\textwidth}
+            %\includegraphics[width=\linewidth]{images/safety_sheet.png}
+    %\end{columns}
+  %\end{frame}
 
-  \begin{frame}{Abstract}
-    Gag algorithms are a staple of computer science and provide many a good chuckle.
+  %\section{Introduction}
+
+  %\begin{frame}{Abstract}
+    %Gag algorithms are a staple of computer science and provide many a good chuckle.
 Take, for instance, MiracleSort\textsuperscript{\color{blue}\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort the list:
-    \begin{center} \texttt{while isSorted(unsorted\_list) is False: \\ \hspace{-70} time.sleep(1000)} \end{center}
-    Despite their apparent uselessness, joke algorithms provide valuable insight into algorithm design and complexity\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}}.
+    %\begin{center} \texttt{while isSorted(unsorted\_list) is False: \\ \hspace{-70} time.sleep(1000)} \end{center}
+    %Despite their apparent uselessness, joke algorithms provide valuable insight into algorithm design and complexity\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}}.
 In this paper we propose an implementation to Quantum Bogosort, one such joke algorithm.
+  %\end{frame}
+
+  %ISEF
+  \begin{frame}{Introduction}
+      Quantum computing is at a critical junction where hardware has finally caught up to decades of theory, which has sparked renewed interest in the technology.
+However, the quantum mechanics which make quantum computers fast also make them difficult to learn how to program.
+      Training the quantum workforce is essential as demand quickly begins to outstrip the throughput of graduate laboratories. \\
+      To meet this need, we propose a quantum algorithm to spark interest in quantum and bridge the understanding between classical and quantum paradigms: \textbf{Quantum bogosort}. \\
+      In the same way that in-jokes like the esoteric programming language ``Ook!" can teach programming concepts, quantum bogosort aims to teach quantum computing fundamentals like superposition and entanglement to beginners while being more accessible than traditional textbooks\textsuperscript{\color{blue}\cite{nielsen_chuang_2021}}.
+      %Its applied and absurd nature make it more engaging than a traditional textbook, %much like an esoteric language (e.g. BrainF**k) can be more interesting than  programming language design
+      %It aims to be more accessible than texts like Chuang's \textit{Quantum Computation and Quantum Information} and more 
   \end{frame}
@@
   \begin{frame}{What is Quantum Bogosort?}
-    \textit{Classical} Bogosort is a sorting algorithm.
+    \textit{Classical} Bogosort\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}} is a (bad) sorting algorithm.
     \begin{enumerate}
@@
     \end{enumerate}
-    This has unbounded worst-case time complexity and an expected time complexity of $O(n \cdot n!)$. \\
+    %This has unbounded worst-case time complexity and an expected time complexity of $O(n \cdot n!)$. \\
     \vspace{10}
-    \textit{Quantum} Bogosort is very similar to classical Bogosort, except it uses a quantum computer to permute the list.
+    \textit{Quantum} Bogosort\textsuperscript{\color{blue}\cite{the_other_tree_2009}} was a proposed extension to bogosort where all list permutations would be placed in superposition.
+With the `many worlds' interpretation of quantum mechanics, any universe not containing the sorted list would be destroyed (this is where the joke departs from reality).
+Using the more popular Copenhagen interpretation, our quantum algorithm acts like a random number generator, selecting one permutation of the list at a time; we chose the latter approach and implemented a novel quantum random number generator (QRNG).
   \end{frame}
 
-  \begin{frame}{Attachments}
-  Code + Paper: \href{https://github.com/Borris-the-real-OG/Quantum_Bogosort}{\color{blue} \underline{github.com/Borris-the-real-OG/Quantum\_Bogosort}}
-  \end{frame}
+  %\begin{frame}{Attachments}
+  %Code + Paper: \href{https://github.com/Borris-the-real-OG/Quantum_Bogosort}{\color{blue} %\underline{github.com/Borris-the-real-OG/Quantum\_Bogosort}}
+  %\end{frame}
 
-  \begin{frame}{Acknowledgements}
-  \begin{itemize}
-  \item \href{https://qiskit.org/}{\color{blue} \underline{Qiskit}}
-  \item \emph{We acknowledge the use of IBM Quantum services for this work.
+  %\begin{frame}{Acknowledgements}
+  %\begin{itemize}
+  %\item \href{https://qiskit.org/}{\color{blue} \underline{Qiskit}}
+  %\item \emph{We acknowledge the use of IBM Quantum services for this work.
 The views expressed are those of the authors, and do not reflect the official policy or position of IBM or the IBM Quantum team.\textsuperscript{\color{blue}\cite{ibm_21}}}
-  \item All code was written and tested with Qiskit 0.32.0 \& Python 3.8 on WSL2 Ubuntu 20.04.1.
-  \item Special thanks to Joe Clapis
-  \end{itemize}
-  \end{frame}
+  %\item All code was written and tested with Qiskit 0.32.0 \& Python 3.8 on WSL2 Ubuntu 20.04.1.
+  %\item Special thanks to Joe Clapis
+  %\end{itemize}
+  %\end{frame}
 
-  \begin{frame}{Purpose}
-  Just like its classical counterpart, Quantum Bogosort is pretty useless.
+  %\begin{frame}{Purpose}
+  %Just like its classical counterpart, Quantum Bogosort is pretty useless.
 In fact, classical sorting algorithms have been proven to have a lower bound time complexity of $ \Omega (N \log_2 N) $: there's nothing to optimize!
 \\
-  \textbf{\textit{However}}, QBS presents a good case study for quantum algorithms; we present an analysis of implementation and complexity of QBS in a similar vein to Gruber et al.'s\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}} analysis of classic bogosort.
-  \end{frame}
+  %\textbf{\textit{However}}, QBS presents a good case study for quantum algorithms; we present an analysis of implementation and complexity of QBS in a similar vein to Gruber et al.'s\textsuperscript{\color{blue}\cite{gruber_holzer_ruepp_2007}} analysis of classic bogosort.
+  %\end{frame}
 
-  \section{Algorithm Design}
+  %\section{Algorithm Design}
 
@@
     \begin{columns}[T,onlytextwidth]
-        \column{0.48\textwidth}
+        \column{0.58\textwidth}
             \vspace{20}
@@
             $$ \sum_{x=0}^{N-1} \frac{1}{\sqrt{N}} \ket{x_{BE}} $$
-            This is quite easy to do with powers of 2; we just need to use Hadamard gates with control qubits to chunk a big number into smaller subproblems.
-       \column{0.48\textwidth}
+            We want an even probability ($\frac{1}{N}$) of measuring any integers from 0 to $N-1$. \\
+            \vspace{1em}
+            Other QRNGs can only generate even superpositions that are powers of 2; our novel QRNG can generate \textit{any} evenly balanced superposition by `chunking' the problem into smaller subproblems.
+            %This is quite easy to do with powers of 2; we just need to use Hadamard gates with control qubits to chunk a big number into smaller subproblems.
+       \column{0.4\textwidth}
+           `Chunking' 0 to 7
            \begin{align*}
@@
     \textbf{Limitation:} Hadamard gates create the state $ \frac{1}{\sqrt{2}} \ket{0} + \frac{1}{\sqrt{2}} \ket{1} $. What about states that aren't powers of 2? \\ \\
-    \textbf{Solution:} Arbitrary rotation gates allow for qubit superposition with arbitrary amplitudes. \\
-    Hadamard is just a rotation of $ \frac{\pi}{2} $ around the Pauli Y-axis!
+    \textbf{Solution:} Arbitrary rotation gates allow for qubit superposition with arbitrary amplitudes.
+    %Hadamard is just a rotation of $ \frac{\pi}{2} $ around the Pauli Y-axis!
     \begin{align*}
         &\ket{\psi} = \ket{0} \\
-        & \text{R}_{\text{y}}(\theta) \ket{\psi} = \begin{bmatrix} \cos \frac{\theta}{2} & -\sin \frac{\theta}{2} \\ \sin \frac{\theta}{2} & \cos \frac{\theta}{2} \end{bmatrix} \ket{\psi} = -\sin \frac{\theta}{2} \ket{0} + \cos \frac{\theta}{2} \ket{1}
+        & \text{R}_{\text{y}}(\theta) \ket{\psi} = \begin{bmatrix} \cos \frac{\theta}{2} & -\sin \frac{\theta}{2} \\ \sin \frac{\theta}{2} & \cos \frac{\theta}{2} \end{bmatrix} \ket{\psi} = %-\sin \frac{\theta}{2} \ket{0} + \cos \frac{\theta}{2} \ket{1}
+        \cos \frac{\theta}{2} \ket{0} + \sin \frac{\theta}{2} \ket{1}
     \end{align*}
-    \textit{Everything in this project is measured in Pauli-Z basis, so we don't worry about phase or imaginary numbers.}
+    \textit{Everything in this project is measured in Pauli-Z basis, the most common computational basis.}%so we don't worry about phase or imaginary numbers.}
   \end{frame}
@@
         \column{0.5\textwidth}
-        $$\theta_0 = 2\arccos \sqrt{\frac{77-\color{violet}64}{77}}$$
+        $$\color{blue}\theta_0\color{black} = 2\arccos \sqrt{\frac{77-\color{violet}64}{77}}$$
         \color{violet} Most Significant Bit in register
@@
     \begin{align*}
-      &\text{R}_y(\theta_0, \psi_0), \ \color{lightgray}\text{X}(\psi_0)\color{black}, \ \text{CH}(\psi_0, \psi_{[1,6]}), \ \text{X}(\psi_0) = \\
+      &\text{R}_y(\color{blue}\theta_0\color{black}, \psi_0), \color{lightgray}\overbrace{\text{X}(\psi_0)}^{\text{Optimized}}\color{black}, \text{CH}(\psi_0, \psi_{[1,6]}), \ \text{X}(\psi_0) = \\
       &\color{green} \frac{1}{\sqrt{77}} (\ket{0000000} + \ket{0000001} + ... + \ket{0111111}) \color{black} + \color{red} \sqrt{\frac{13}{77}} \ket{1000000}
@@
 
-  \section{Practicality Assessment/Analysis}
+  %\section{Practicality Assessment/Analysis}
 
-  \begin{frame}{Circuit}
+  \begin{frame}{Generated Circuit}
     \begin{figure}
@@
            \includegraphics[width=\linewidth]{images/example_measurement.png}
-           \textit{Example measurement with }\texttt{parts = 15}. \\
+           \textit{Example measurement with } $N = 15$. \\%\texttt{parts = 15}. \\
            \end{center}
            \textbf{Complexity} \\
-           $n =$ \# elements, $o =$ \texttt{bin(n).count(`1')} \\
+           $n =$ \# elements \\
+           $o =$ \texttt{bin(n).count(`1')} \\
            Many-worlds: $O(o)$ \\
@@
            \column{0.48\textwidth}
-           Gate count scales with the number of `1's in the binary representation, \textit{\textbf{not}} the size of the number.
+           Gate count scales with the number of ones in $N$'s binary representation, \textit{\textbf{not}} $N$ itself.
            \includegraphics[width=\linewidth]{images/resources.png}
-           \textit{Circuit depth, width, and total gate count in relation to} $ i $.
+           \textit{Circuit depth, width, and total gate count in relation to} $N$.
+       \end{columns}
+   \end{frame}
+
+   \begin{frame}{Addressing Error}
+    \begin{columns}
+       \column{0.4\textwidth}
+       \small \null \\
+       \vspace{10pt}
+        Why don't we use $\chi^2$ to check our results?
+We can calculate the evolution of a qubit's state through statevectors or density matrices (e.g.
+$\rho$) to prove the probability amplitudes are correct.
+\\
+       \textbf{However}, today's quantum computers are noisy --- real hardware doesn't perform under these ideal conditions.
+$\rho'$ shows how the state purity changes once the circuit is transpiled with error added to the gates.
+       \column{0.6\textwidth}
+       \begin{split*}
+       &\rho = \underbrace{
+       \begin{bNiceArray}{>{\strut}ccccc}[margin,extra-margin = 1pt]
+        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
+        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
+        \vdots & \vdots & \ddots & \vdots & \vdots \\
+        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
+        0 & 0 & \cdots & 0 & 0 \\
+        \CodeAfter
+        \begin{tikzpicture}
+         \node [draw=red, rounded corners=6pt, inner ysep = 1pt,
+            rotate fit=-39, fit = (1-1) (5-5) ] {} ;
+        \end{tikzpicture}
+       \end{bNiceArray}}_{16 \times 16, \ \color{red}\text{Purity} \ = \ 1} \\[10pt]
+       &\rho' = \underbrace{
+       \scalemath{0.35}{
+       \begin{bNiceArray}{>{\strut}ccccc}[margin,extra-margin = 1pt]
+        0.06997 & 0.04275 - 0.00057i & \cdots & 0.0376 + 0.00028i & 0.00036 + 0.00024i  \\
+        0.04275 + 0.00057i & 0.06646 & \cdots & 0.03687 + 0.00025i & 0.00002 + 0.0001i  \\
+        \vdots & \vdots & \ddots & \vdots & \vdots \\
+        0.0376 - 0.00028i & 0.03687 - 0.00025i & \cdots & 0.06272 & 0.00217 + 0.00045i  \\
+        0.00036 - 0.00024i & 0.00002 - 0.0001i & \cdots & 0.00217 - 0.00045i & 0.01268  \\
+        \CodeAfter
+        \begin{tikzpicture}
+         \node [draw=red!50, rounded corners=4pt, inner ysep = -1pt,
+            rotate fit=-11, fit = (1-1) (5-5) ] {} ;
+        \end{tikzpicture}
+       \end{bNiceArray}
+       }}_{16 \times 16, \ \color{red}\text{Purity} \ = \ 0.42118 \\}
+       \end{split*}
+       \tiny \centering \textit{Using noise model from IBMQ\textsuperscript{\color{blue}\cite{ibm_21}} Santiago.
+Note that this does not take into account measurement, reset, relaxation or decoherence errors.} % It could, but I'm lazy.
+¯\_(ツ)_/¯
    \end{columns}
@@
   \begin{frame}{Conclusion}
-  Quantum Bogosort is not a useful algorithm, but it's helpful in teaching quantum computing fundamentals in a fun way. \\ \\
-  Although infeasible due to its ballooning gate count, the QRNG algorithm could be repurposed for quickly preparing evenly balanced states among qubits. \\ \\
-  By trying to optimize around the abysmal $ O(\log_2 n!) $ space complexity via `dividing and conquering', we end up implementing quicksort.
+  %Quantum Bogosort is not a practical algorithm, but it's helpful in teaching quantum computing fundamentals in a fun way. \\
+  Although potetially infeasible due to its ballooning gate count, the QRNG algorithm could be repurposed for state preparation in tandem with other quantum algorithms.
+However, it's not necessarily the \textit{fastest} way for state preparation; transpilation optimizations and optimal gate decompositions pose interesting extensions to this problem that are especially pertinent with current quantum computers.
+Additionally, it could be interesting to further design around the constraints of NISQ devices by executing QRNG incrementally in order to reduce decoherence and relaxation errors.
+\\%Although this is one way to achieve even measurement amplitudes over a register, it's not necessarily the \textit{fastest} way; transpilation optimizations and optimal gate decompositions pose interesting extensions to this problem that are especially pertinent with current quantum computers.
+\\
+  %There are dozens of error mitigation schemes for quantum computers, however implementing robustness was not a focus of this project.
+For example, by trying to optimize around the abysmal $ O(\log_2 n!) $ space complexity via `dividing and conquering', we end up implementing quicksort!
+%Quantum Bogosort doesn't provide a quantum advantage over its classical counterpart (except by being truly random)
+  \vspace{0.5em}
+  Although we successfully implemented quantum bogosort and made certain quantum computing concepts more accessible, this algorithm did not cover interference/phase --- important, fundamental ideas that are present throughout most of quantum computing.
   \end{frame}

### 2023-03-27 paper/main.tex
late for sigbovik? never!

@@
 \usepackage{hypcap}
-\usepackage{fancyhdr}
+\usepackage{authblk}
 
@@
 
-\doublespacing
+\pagenumbering{gobble}
+
+\newlength{\bibitemsep}\setlength{\bibitemsep}{.2\baselineskip plus .05\baselineskip minus .05\baselineskip}
+\newlength{\bibparskip}\setlength{\bibparskip}{0pt}
+\let\oldthebibliography\thebibliography
+\renewcommand\thebibliography[1]{%
+  \oldthebibliography{#1}%
+  \setlength{\parskip}{\bibitemsep}%
+  \setlength{\itemsep}{\bibparskip}%
+}
 
 \title{Quantum Bogosort}
-\author{George Huebner}% (Grade 12), Michael Caines (Teacher sponsor)}
-\date{May 7, 2022}
+\author[1]{George Huebner}
+\affil[1]{University of Illinois at Urbana-Champaign}
+\author[2]{Beam Fujimoto}
+\affil[2]{Inter IKEA Holding B.V.}
+\date{April 0, 2023}
 
 \begin{document}
-\renewcommand{\abstractname}{Preface}
-\pagenumbering{gobble}
+\renewcommand\refname{\centering \line(1,0){250}}
+\renewcommand{\abstractname}{Abstract}
 
-\begin{titlepage}
 \maketitle
-\centering
-\large Walter Payton College Prep, Chicago IL, United States \\
-ISEF Systems Software/Algorithms (SOFT055)% \\
-%\large Illinois STEM Fair, Computer Science Research, Design Category \\
-%\vspace{3em}
-%\huge Sponsor \\
-%\vspace{1em}
-%\large Michael Caines \hspace{15em}
-\end{titlepage}
-
-\pagestyle{fancy}
-\fancyhead[R]{Huebner: \textit{Quantum Bogosort} (SOFT055)}
-\fancyhead[L]{}
 
@@
 \hypersetup{hidelinks}
-\tableofcontents
 }
-\pagebreak
-
-\pagenumbering{arabic}
 
@@
 
-Classic Bogosort is a joke sorting algorithm where a list is randomly permuted and checked to see if it is sorted. In pseudocode:
+Classic Bogosort is a sorting algorithm where a list is randomly permuted and checked to see if it is sorted. In pseudocode:
 
@@
 Quantum Bogosort\textsuperscript{\cite{the_other_tree_2009}} (QBS), then, is a variation on classical bogosort.
-Based on the \href{https://www.pbs.org/wgbh/nova/manyworlds/pdf/dissertation.pdf}{many worlds} interpretation of quantum mechanics, QBS proposes that a quantum computer can superimpose all possible permutations of a list and destroy any universe in which a sorted list does not exist, thus sorting the list in $ O(1) $ time.
+Based on the \href{https://www.pbs.org/wgbh/nova/manyworlds/pdf/dissertation.pdf}{many worlds} interpretation of quantum mechanics, QBS proposes that a quantum computer can superimpose all possible permutations of a list and destroy all universes in which the list is not sorted, thus sorting the list in $ O(1) $ time.
 Because parallel universes are non-communicating, there's no way to know the list is sorted other than by the principle of quantum necessity.
-The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) uses the same algorithm, but the quantum circuit must be run until the list is sorted.
+The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) uses the same algorithm, but the quantum circuit must be run until the list is observed to be sorted.
 
@@
 \section{Background}
-Most people are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to develop and test QBS.
-This is not meant to be a complete explaination of the quantum mechanics that underlie this project, though, so let the physics enthusiast proceed carefully.
+Most people (authors included) are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to develop and test QBS.
+This is not meant to be a complete explanation of the quantum mechanics that underlie this project, though, so let the physics enthusiast proceed carefully.
 
 \subsection{Quantum Computers}
-Quantum computers should be thought of not as processors that parse instructions, but rather circuits composed of gates that operate on a group of qubits (\textbf{qu}antum \textbf{bit}, the quantum unit of information).
-While revolutionary, quantum computers only achieve speedups in certain applications, and have not yet cleared a benchmark of true quantum supremacy due to current hardware implementations.
-Current noisy intermediate scale quantum (NISQ) computers are highly error prone and have comparatively few qubits and are better thought of as a separate processing unit to be orchestrated by a classical computer.
+The reader is best served by thinking of quantum computers not as Turing machines (although, naturally, quantum computers \textit{are} Turing machines), but rather circuits composed of gates that operate on registers of qubits (\textbf{qu}antum \textbf{bit}, the quantum unit of information).
+Current noisy intermediate scale quantum (NISQ) computers are highly error prone and have comparatively few qubits, which proves challenging for researchers to prove increasingly contrived claims of quantum supremacy.
 \subsection{Qubits}
-Qubits have quantum properties of superposition, entanglement, and interference that distinguish them from classical bits.
-Whereas a classical bit is either 1 or 0, a qubit is in a superposition of 1 and 0, which are defined as the following kets:
-$$ \ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \text{ (Classical 0)} $$
-$$ \ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix} \text{ (Classical 1)} $$
-Qubit superpositon can be thought of as having a certain amount (amplitude) of `one-ness' and `zero-ness' that can be represented as so:
+Qubits have quantum mechanical properties of superposition, entanglement, and interference that distinguish them from classical bits.
+Whereas a classical bit is either 1 or 0, a qubit is in a superposition of the $\ket{1}$ and $\ket{0}$ eigenstates.
+To actually make use of qubits, they must be measured, which collapses them to either $\ket{0}$ or $\ket{1}$.
+This is a probabilistic event, with the quantum mechanical measurement rule determining the probabilities as follows:
 $$ \ket{x} = \alpha \ket{0} + \beta \ket{1} \ | \ \text{ Probability}_{\ket{0}} = | \alpha |^2 \text{, } \text{ Probability}_{\ket{1}} = | \beta |^2 $$
-%Note that $|x|$ here represents magnitude as $\alpha$ and $\beta$ are imaginary.
-To actually make use of qubits, they must be measured, which collapses them to either $\ket{0}$ or $\ket{1}$.
-This is a probabilistic, quantum event, hence the probability\textsubscript{$\ket{x}$}.
 It is helpful to consider the \hyperref[fig:bsphere]{Bloch sphere} when analyzing qubit state and rotation around the Pauli axes; essentially, the `North Pole' is defined as $\ket{0}$ and the `South Pole' is defined as $\ket{1}$.
@@
 $$ \text{Z}(\ket{1}) = -\ket{1} $$
-$$ \text{H} = \text{R}_{\text{y}} (\frac{\pi}{2}) \text{Z} \ \ \ \text{(Hadamard gate)}$$
+$$ \text{H} = \text{R}_{\text{y}} \left(\frac{\pi}{2}\right) \text{Z} \ \ \ \text{(Hadamard gate)}$$
 $$ \text{H}(\ket{0}) = \frac{1}{\sqrt{2}} \ket{0} + \frac{1}{\sqrt{2}} \ket{1} = \ket{+} $$
@@
 Notice that R\textsubscript{y} is an arbitrary rotation around the Pauli Y-axis; Similarly, R\textsubscript{x} and R\textsubscript{z} allow arbitrary rotations along the X- and Z-axes, respectively.\\
-Multiple qubits can be used together for entanglement and amplitude interference, and a group of qubits is called a \textit{register}.
-Although a register can be simplified into a single tensor by taking the tensor product of the qubits in the register, keeping registers in kets makes the math much simpler.
-Bra-ket (Dirac) notation is used as Kronecker products quickly become unwieldy when dealing with large qubit registers.
-
-\begin{figure}[h]
-    \centering
-    $$ \ket{x} \otimes \ket{y} = \begin{bmatrix} x_0 \\ ..
-\\ x_n \end{bmatrix} \otimes \begin{bmatrix} y_0 \\ ..
-\\ y_m \end{bmatrix} = \begin{bmatrix} x_0 \cdot \begin{bmatrix} y_0 \\ ..
-\\ y_m \end{bmatrix} \\ ..
-\\ x_n \cdot \begin{bmatrix} y_0 \\ ..
-\\ y_m \end{bmatrix} \end{bmatrix} = \ket{xy} $$
-    \caption{Elegance of Bra-ket notation}
-    \label{fig:my_label}
-\end{figure}
+Multiple qubits can be used together for entanglement and amplitude interference --- a group of qubits is called a \textit{register}, and is simply a tensor product of discrete quantum states.
 
 \subsection{Entanglement}
-Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an `if' clause in a classical programming language).
-This allows for multiple qubits to be entangled with each other, essentially tying the 'fate' of the qubits together so that they can no longer be thought of as individual qubits but rather a single system.
+Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an `if' statement in a classical programming language), which allows for multiple qubits to be entangled with each other.
 Measuring one qubit will also collapse the other entangled qubits, even if they were never observed.
@@
 The implementation will be written using Qiskit, a Python framework for creating, running and simulating quantum circuits, and tested with the IBMQ Aer and Manila simulators.
-Let us define the following:
+We define the following:
 
@@
 \end{minted}
-\textit{Note that Python numbers have dynamic size, but our 8-bit size would afford us an unsigned char, that is:}
+\textit{Python numbers have dynamic size, but our 8-bit size would afford us an unsigned char, that is:}
 $ x \in \mathbb{Z} \mid 0 \leq x < 256 $ \\
@@
 Instead of storing the list, we create $N = l!$ evenly balanced register states ($l!$ being the number of permutations of $array$).
-A set of states $[0, N) \cap \mathbb{N}$ will be prepared, each with an even amplitude.
+A set of states $[0, N) \subset \mathbb{N}$ will be prepared, with each state having an equal amplitude.
 Concisely:
@@
 
-Because we're storing our number in binary, we want the power of 2 greater than or equal to $l!$.
+Because we're storing our number in binary, we want the power of $2$ greater than or equal to $l!$.
 This means we have $O(\lceil \log_2 l!
-\rceil)$ space complexity, which seems scary at first.
-However, with present-day NISQ technology, this is significantly more favorable than $O(l \cdot s)$ for any useful data type.
-Also keep in mind that because element size isn't a concern, any size element or even elements of different types could be stored in the list (as long as every type used can be compared to each other).
+\rceil)$ space complexity; with present-day NISQ technology, this is significantly more favorable than $O(l \cdot s)$ for any useful data type.
+Also keep in mind that because element size isn't a concern, elements can be any type with a partial ordering.
 
@@
 Before we proceed, we would like to discuss the naive approach to quantum RNG.
-Applying H\textsubscript{ALL} (apply Hadamard to all qubits) to the entire register can create a number of superimposed states $ n $ such that $ n $ is a power of 2.
+Applying H\textsuperscript{$\otimes n$} (apply Hadamard to all qubits) to the entire register can create a number of superimposed states $ n $ such that $ n $ is a power of 2.
 We could still use this output to generate a random number $ h $ such that $ h \in [i, j] $:
@@
 \section{Quantum Bogosort}
-\subsection{Quantum Random Number Generation (QRNG) Algorithm}
-The main algorithm is recursive, essentially moving through the register from most to least significant qubit, performing a R\textsubscript{y} gate to split up a qubit into P 'parts,' and entangling the remaining qubits with controlled Hadamard (CH) gates.
+\subsection{State Preparation (SP) Algorithm}
+The main algorithm is recursive, essentially moving through the register from most to least significant qubit, performing a R\textsubscript{y} gate to split up a qubit into P `parts,' and entangling the remaining qubits with controlled Hadamard (CH) gates.
 \\
-Let's use $i = 77$ as an example.
+\indent Let's use $i = 77$ as an example.
 Recall that for storing an integer $i$, the most significant bit (MSB) used is $\lfloor \log_2 i \rfloor$: For example, the MSB of 77 would be $\lfloor \log_2 77 \rfloor = 64$.
-Because QRNG generates the integers $[0, 77)$ (excluding 77 itself in order to generate 77 integers), we know that $[0, 64)$ is also being generated, because $[0, 64) \subseteq [0, 77)$.
+Because SP generates the integers $[0, 77)$ (excluding 77 itself in order to generate 77 integers), we know that $[0, 64)$ is also being generated, because $[0, 64) \subseteq [0, 77)$.
 \\
-This means that every possible state of the qubit register with a leading zero (in this case, in the 64s place) should be generated.
+\indent This means that every possible state of the qubit register with a leading zero (in this case, in the 64s place) should be generated.
 That's already really easy to do!
@@
 Also note the X gate applied to the MSB because we really want an anti-controlled, not a normal controlled gate.
+The problem is that we've unevenly split the integer 77 into 2: $|[0, 64)| \neq |[64, 77)|$.
+Numbers in the range $[64, 77)$ have a $\left(\frac{1}{\sqrt{13}} \cdot \frac{1}{\sqrt{2}}\right)^2$ probability of being measured whereas numbers in the $[0, 64)$ range have a $\left(\frac{1}{\sqrt{64}} \cdot \frac{1}{\sqrt{2}}\right)^2$ probability of being measured.
 \\
-The problem is that we've unevenly split the integer 77 into 2: $|[0, 64)| \neq |[64, 77)|$.
-Numbers in the range $[64, 77)$ have a $(\frac{1}{\sqrt{13}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured whereas numbers in the $[0, 64)$ range have a $(\frac{1}{\sqrt{64}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured.
+
+\noindent To fix this, we have to change the $\frac{1}{\sqrt{2}}$ from the Hadamard gate to something that gives us a more complex superposition; to do this we use the arbitrary rotation gates.
+X and Y gates give us the desired rotations we want, but we chose Y so we don't have to worry about phase (the sign of a qubit).
+Phase doesn't matter in this case because only the magnitudes effect the probability of measurement, but it might be important if using SP in another algorithm.
 \\
-To fix this, we have to change the $\frac{1}{\sqrt{2}}$ from the Hadamard gate to something that gives us a more complex superposition; to do this we use the arbitrary rotation gates.
-$X$ and $Y$ gates give us the desired rotations we want, but we chose $Y$ so we don't have to worry about phase (the sign of a qubit).
-Phase doesn't matter in this case because only the magnitudes effect the probability of measurement, but it might be important if using QRNG in another algorithm.
+\indent Superpositions can be represented using spherical coordinates.
+For example, on the \hyperref[fig:bsphere]{Bloch sphere}, the spherical coordinates $(1, 0.955, 0)$ would represent the state $\sqrt{\frac{1}{3}} \ket{0} + \sqrt{\frac{2}{3}} \ket{1}$ because $\cos (0.955) \approx \sqrt{\frac{1}{3}}$.
+We can do some algebra to get a $\theta$ for an arbitrary superposition.
+For SP, this looks like $\theta = 2\arccos\left(\sqrt{\frac{77 - MSB}{77}}\right)$.
 \\
-Superpositions can be represented using spherical coordinates.
-For example, on the \hyperref[fig:bsphere2]{Bloch sphere}, the spherical coordinates $(1, 0.955, 0)$ would represent the state $\sqrt{\frac{1}{3}} \ket{0} + \sqrt{\frac{2}{3}} \ket{1}$ because $\cos (0.955) \approx \sqrt{\frac{1}{3}}$.
-We can do some algebra to get a $\theta$ for an arbitrary superposition.
-For QRNG, this looks like $\theta = 2\arccos(\sqrt{\frac{77 - MSB}{77}})$.
-\\
-\begin{figure}[H]
-    \centering
-    \capstart
-    \includegraphics[width=0.19\textwidth]{images/bloch_sphere2.png}
-    \caption{Bloch sphere\textsuperscript{\cite{wikipedia_2021}}}
-    \label{fig:bsphere2}
-\end{figure}
-\noindent Recall that we have to apply a $X$ gate because gates are controlled by $\ket{1}$, not $\ket{0}$.
-We can save an extra gate by doubling our rotation (hence the 2), which gives us $\sqrt{\frac{77 - MSB}{77}} \ket{0}$ instead, which flips right back to $\ket{1}$ with another $X$ gate.
+\indent Recall that we have to apply an X gate because gates are controlled by $\ket{1}$, not $\ket{0}$.
+We can save an extra gate by doubling our rotation (hence the 2), which gives us $\sqrt{\frac{77 - MSB}{77}} \ket{0}$ instead, which flips right back to $\ket{1}$ with another X gate.
 \\
@@
 
-\noindent The differences between what is conceptually happening and what is actually implemented are shown below.
-$\bullet$ is a control and $\circ$ is an anti-control (controls on $\ket{0}$ instead of $\ket{1}$).
-Note that H\textsubscript{ALL} is \textbf{not} a multi-qubit gate --- it's illustrated that way for clarity.
+\noindent The conceptual circuit diagram is shown below; note that $\bullet$ is a control and $\circ$ is an anti-control (controls on $\ket{0}$ instead of $\ket{1}$), and H\textsubscript{ALL} is syntactic sugar for a singly-controlled H\textsuperscript{$\otimes n$}.
 \begin{figure}[H]
@@
     \caption{Conceptual circuit\textsuperscript{\cite{chuang}}}
-    \vspace{10mm}
-    \includegraphics[width=\textwidth]{images/circuits/optimized_circuit.pdf}
-    \caption{Circuit with proper controls\textsuperscript{\cite{chuang}}}
     \label{fig:circs}
@@
 \begin{itemize}
-    \item Multi-controlled Hadamard (MCH) gates are implemented manually with an ancilla qubit and multi-controlled X (MCX) gates.
+    \item Multi-controlled Hadamard (MCH) gates are implemented manually with an ancilla qubit and Toffoli (multi-controlled X) gates.
 This could almost certainly be optimized more on a case-by-case basis (i.e.
@@
 circuit.measure(register, measurement)
-
-count = execute(circuit, Aer.get_backend("aer_simulator"), shots=1024) //
-.result().get_counts(circuit)
-plot_histogram({key[::-1]: value for key, value in count.items()})
 \end{minted}
@@
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
@@
 \subsection{Practicality Assessment}
-Obviously bogosort is not meant to be a practical algorithm, but nevertheless we will conduct a review of the efficiency of QRNG.
+Obviously bogosort is not meant to be a practical algorithm, but nevertheless we will conduct a review of its performance.
 \subsubsection{Complexity}
@@
 \\
-For the integer $i$ passed into the QRNG algorithm, we allocate $\lceil \log_2 i \rceil + 1$ qubits.
+For the integer $i$ passed into the SP algorithm, we allocate $\lceil \log_2 i \rceil + 1$ qubits.
 However, this algorithm is rather unique in that gate count and depth don't scale exactly with $i$, but rather \mintinline{python}{bin(i).count('1')}.
@@
 \\
-Consequently, there are large drops in complexity around powers of 2:
+\indent Consequently, there are large drops in complexity around powers of 2:
 \begin{align*}
@@
 Of course, the main problem with this ballooning circuit size is the fidelity of the measurement; we work with imperfect machines and error rates can compound over time to make a measurement meaningless.
-That being said, this is a QRNG algorithm, so the noise and error generated don't make much of a difference in this scenario (given that the noise is truly random and evenly distributed) besides occasionally measuring a number $ \geq i$.
+That being said, this is by nature a randomized algorithm, so the noise and error generated don't make much of a difference in this scenario (given that the noise is truly random and evenly distributed) besides occasionally measuring a number $ \geq i$.
 \\
@@
 This would allow for the approximation of the expected position on the Bloch sphere during measurement, thus providing a more accurate error rate.
-One way we can model error is by wrapping our gates with error gates --- identity gates with an error rate corresponding to a noise profile sampled from real quantum hardware.
+\indent One way we can model error is by wrapping our gates with error gates --- identity gates with an error rate corresponding to a noise profile sampled from real quantum hardware.
 This approach works with gate errors, including measurement and reset errors, and could also be applied to decoherence and relaxation errors.
 \\
-Often times noise on hardware will turn a pure quantum state into a mixed state, a composition of pure states with classical probabilities (which is \textbf{not} the same as superposition); for example, perhaps an R\textsubscript{Y}$(\frac{\pi}{2})$ will succeed $80\%$ of the time and over/undershoot by $\frac{\pi}{6}$ 20\% of the time.
+\indent Often times noise on hardware will turn a pure quantum state into a mixed state, a composition of pure states with classical probabilities (which is \textbf{not} the same as superposition); for example, perhaps an R\textsubscript{y}$\left(\frac{\pi}{2}\right)$ will succeed $80\%$ of the time and over/undershoot by $\frac{\pi}{6}$ 20\% of the time.
 We can write this mixed state as
@@
 \end{gather*}
+\subsection{Comparison to Other State Preparation Algorithms}
+Araujo et al.\textsuperscript{\cite{araujo}} use a very similar divide and conquer strategy to recursively calculate the angles for R\textsubscript{y} gates.
+Although our algorithm was developed independently, it could be seen as a special case of Araujo's algorithm by exploiting the structure of an interval of binary integers to use H instead of arbitrary rotation gates.
+\\
 
+\noindent Zhang et al.\textsuperscript{\cite{zhang}} achieve arbitrary state preparation with an impressive depth of $\Theta(n)$, at the cost of $O(2^n)$ additional ancillary qubits.
+The ``parallelization'' of CNOT gates employed in their algorithm is also applicable to SP.
+\\
+
+\noindent M\"ott\"onen et al.\textsuperscript{\cite{mottonen}} describe an arbitrary state preparation algorithm from $\ket{a}$ to $\ket{b}$.
+Unlike M\"ott\"onen state preparation, SP starts from $\ket{0}^{\otimes n}$ and completely disregards phase because only the computation basis is considered.
 \subsection{Bogosort}
@@
     \item Generate all permutations of list with length $l$ (symbolic representations are fine to avoid memory overhead)
-    \item Use QRNG with parts = $l!$
+    \item Use SP with parts = $l!$
     \item Using the measured value from step 2, randomly select a permutation of the list
@@
 \end{enumerate}
-\vspace{15mm}
 Note that with the many-worlds interpretation we can stop at step 4, because a universe where the list is sorted must exist (assuming the list can be sorted) by quantum necessity.
@@
 \subsection{Complete Algorithm}
-Using QRNG algorithm defined in \ref{qrng}
+Using the algorithm defined in \ref{qrng}
 \begin{minted}{python}
@@
 
-\pagebreak
-
 \section{Conclusion}
-In this paper, we implemented the quantum bogosort algorithm using a novel QRNG to randomly permute the list.
-Although QBS is a joke, the QRNG algorithm implemented could potentially be useful in setting states where an H\textsubscript{ALL} wouldn't do the trick.
+In this paper, we implemented the quantum bogosort algorithm by using a novel state preparation routine to randomly permute the list.
+Although QBS is a joke, the state preparation algorithm implemented could potentially be useful in setting states where H\textsubscript{ALL} wouldn't do the trick.
 \\
-The QRNG algorithm presented could be further studied and improved for efficient preparation of arbitrarily superimposed qubits.
-Right now, QRNG generates an interval of integers, but this could be extended to generating an even superposition of arbitrary values.
-Additionally, as discussed in the practicality assessment, this algorithm does not scale well due to the prevalence of multi-controlled gates.
+Right now, the SP routine generates an interval of integers, but this could be extended to generating an even superposition of arbitrary values.
+As discussed in the practicality assessment, this algorithm does not scale well due to the prevalence of multi-controlled gates.
 \\
 Another way to improve QBS would be to study how the list is sorted.
-Because the (non-unique) permutations of a list grows at $n!$, QRNG's $O(\log_2 n!)$ space complexity quickly becomes overwhelmed with larger lists.
+Because the (non-unique) permutations of a list grows at $n!$, SP's $O(\log_2 n!)$ space complexity quickly becomes overwhelmed with larger lists.
 Finding ways to reduce this inefficiency would be ideal, perhaps by using a `divide and conquer' strategy (\`{a} la quicksort); However, this would defeat the purpose of bogosort!
@@
 \\
-%We were able to both exercise previous knowledge of quantum computing and learn new skills in data collection and presentation.
-Additionally, we spent more time designing this algorithm because there's no previous implementation of QBS to compare against: Consequently we got to pursue many different solutions and choose the most interesting one.
-\\
 
-\vspace{10mm}
-\noindent All code was written and tested with Qiskit 0.32.0 \& Python 3.8 on WSL2 Ubuntu 20.04.1.
+\section*{\small \centerline{Acknowledgements}}
 
-\vspace{15mm}
+\noindent All code was written and tested with Qiskit 0.32.0 and Python 3.8 on WSL2 Ubuntu 20.04.1. \\
+
 \noindent \emph{We acknowledge the use of IBM Quantum services for this work.
 The views expressed are those of the authors, and do not reflect the official policy or position of IBM or the IBM Quantum team.\textsuperscript{\cite{ibm_21}}}
-
-\pagebreak
-
+\vspace{-1.5em}
 \bibliographystyle{ieeetr}
 \nocite{*}
-\bibliography{sources}
+\small{\bibliography{sources}}
 
